---
title: "R Notebook"
date: "`r paste('Last edits on:',date(), sep=' ')`"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document: default
  html_notebook: default
---

```{r preamble, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	background = "#ADD8E6",
	cache = TRUE,
	tidy = 'styler',
	class.source = "bg-warning",
	class.output = "bg-success",
	fig_caption = TRUE,
	digits = 3
)


```

```{r libraries, include=FALSE}
using<-function(...) {
    libs<-unlist(list(...))
    req<-unlist(lapply(libs,require,character.only=TRUE))
    need<-libs[req==FALSE]
    if(length(need)>0){ 
        install.packages(need)
        lapply(need,require,character.only=TRUE)
    }
}
using("tidyverse","ggplot2","psych","corrplot","DT","formatR","ggExtra","car","rethinking","rstan","rvest","dagitty","priceR","lubridate")

round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
  df[,nums] <- round(df[,nums], digits = digits)
  (df)
}
```

# Data

I need FOREX data captured by day (end of day) for consecutive days prior to and throughout the Russian war on Ukraine.  Before getting into to many details, I need not just the Ruble data but I capture as many relevant variables during that time to better understand (and predict) subsequent Ruble exchange values (relative to the US dollar).

```{r RubleData}
dat.ex <- priceR::historical_exchange_rates("USD","RUB","2022-02-01",paste(lubridate::today()))
dat.ex
```





```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
