---
title: "Patrick's Class Notes"
output: html_notebook
---

```{r preamble, message=FALSE, warning=FALSE}
library(rethinking)
```


# Chapter 1

Prepare yourself to learn statistics from a different angle.  Also, get ready to learn some skills (and tools).  Those include R, Rstudio, stan, and for the daring, tidyverse.

## Getting started

1. Install R - get the latest version for your platform
2. Install RStudio - get the **preview** version instead of the standard download.
3. Install rstan - see [mc-stan.org](https://mc-stan.org) for details

## Getting prepared each week

1. Read the assigned book chapter.  I assume all graduate students read so please do so.
2. Watch the videos.  If you find an additional loop through the readings helps, I strongly encourage you to watch them.  Richard does a fine job providing further details from his book.
3. Come prepared to ask questions on Tuesdays.
4. Come prepared to do some homework together - Tuesdays (occasionally) and Thursdays (always).
5. Stay ahead of the workflow.  
6. Check for additional references including the [R code from the book](http://xcelab.net/rmpubs/sr2/code.txt) and the efforts to port that code to more modern tools.  Please consult [Richard's webpage](https://xcelab.net/rm/statistical-rethinking/) for more links.

# Chapter 2

Probability theory requires some work.  We will work on a problem together and then solve the chapter problems together on Thursday.  Here are the problems I want us to work on together during the Tuesday lecture period:

## Some fundamentals

[Look here](https://www.statlect.com/) for some useful information during the course.

1. Definitions:
  - Probability refers to an event (i.e., something happens): $P(A)$
  - Probability is a number (but it can be qualified later as a qualitative feel) and ranges between 0 and 1:  $0 \le P(A) \le 1$
  - Rules define how we compute and evaluate probabilities
2. Probabilities are ratios of counts...
  - An event occurring count (Numerator)
  - Total count of observations (Denominator)
  - Thus: $p = Event_{count} / (Event_{count} + Non-Event_{count})$
3. Arithmetic on probabilities are easy when...
  - Events are independent
  - Total count of observations are readily available
4. Rules of probability (you will learn these implicitly over time)
  - Addition Rule: $$P(A or B) = P(A \cup B) = P(A) + P(B) - P(A \cap B)$$
  - Complementary Rule: $$P(A) + P(A^{\prime}) = 1$$
  - Conditional Rule: $$P(B, given A) = P(B | A)=P(A \cap B) / P(A)$$
  - Multiplication Rule: $$P(A and B) = P(A \cap B) = P(A) \cdot P(B|A) $$

## An Example

Let's attack psychological science since we are all in the game now.  We appear to be in a crisis where most if not all our published results are either being retracted or questioned based upon glaring failures to replicate.  If a study fails to replicate, should we be surprised?  And, would this surprise lead us to feel as if we are in crisis?  Here are some numbers:

1. Assumptions
  - we are all frequentists and employ some form of NHST (e.g., ANOVA, MRC, etc.).
  - we use a simple probability of 5% error to define "significance" (i.e., $p \le 0.05$).
  - Expected Base Rate (priors condition):  we only venture where it is safe to venture ($P(H_{ours}) \ge P(H_{alts})$)
  - Evidence Discovery Rate:  we only test what is readily available to us ($P(D_{ours}) \ge .5$)
  - True Positive Rate:  we measure with almost perfect validity and precision (a point I shall elaborate upon) but for now, we specify it as a probability ($P(Reject H_0 | H_0 = F) = .99$, aka statistical power or, for our question though...$P(D_{ours} = T | H_{ours}=T) = .99$)
  - False Positive Rate:  we have a very low false positive rate ($P(D_{ours} = T| H_{ours} = F) \le .01$)
  - every study is independent of every other study - including our own
  - subjects are replicates of one another so sample-to-sample variance is zero
  - Base Rate for $H_{ours}$: there are 100 hypotheses available for the data in question and ours is only one of the 100 including the null ($P(H_{ours}) \approx .02$ or double for each $P(H_{alt}$)
2. Computations
  - Use Bayes Theorem with limited information
  $$P(H_{ours}|D) = \frac{P(D_{ours}|H_{ours}) \cdot P(H_{ours}))}{P(D)}$$
  $$P(H_{ours}|D) = \frac{.99 * .02}{.5} = 0.0396 \approx 4\%$$
  - Alternative with full information
  $$P(H_{ours}|D) = \frac{P(D_{ours}|H_{ours}=T) \cdot P(H_{ours}))}{P(D_{ours}|H_{ours}) \cdot P(H_{ours}) + P(D_{ours}| H_{ours} = F) \cdot (1 - P(H_{ours}))}$$
  $$P(H_{ours}|D) = \frac{.99 * .02}{(.99 * .02)+(.01 * .98)} = 0.669 \approx 67\%$$
  - Why the discrepency?
  Consider Richard's example.  He used $P(D_{ours} | H_{ours} = T) = 0.95$, $P(D_{ours} | H_{ours} = F) = 0.05$, $P(H_{ours}) = 0.01$, gives us:
  $$P(H_{ours}|D) = \frac{0.95 * 0.01}{(0.95 * 0.01)+(0.05 * 0.99)} = 0.161 \approx 16\%$$
  - Use your own numbers and see how well you can make the replication crisis seem worse or better.

## Assigned Problems:

### All Easy ones (hints below)
  - 2E1: Notation and Definitions
  - 2E2: Notation
  - 2E3: Notation and Definitions
  - 2E4: Definitions and Intentions of Probability

### Medium ones:
  - 2M1: Use the following code (revealed after class) to produce the following:
```{r 2M1, echo=TRUE, message=FALSE, warning=FALSE}
p_grid <- seq( from=0 , to=1 , length.out=100 )
# likelihood of 3 water in 3 tosses
likelihood1 <- dbinom( 3 , size=3 , prob=p_grid ) # note this is for (1)
likelihood2 <- dbinom( 3 , size=4 , prob=p_grid ) # note this is for (2)
likelihood3 <- dbinom( 5 , size=7 , prob=p_grid ) # note this is for (3)
prior <- rep(1,100) # uniform prior
posterior1 <- likelihood1 * prior
posterior1 <- posterior1 / sum(posterior1) # standardize
posterior2 <- likelihood2 * prior
posterior2 <- posterior2 / sum(posterior2) # standardize
posterior3 <- likelihood3 * prior
posterior3 <- posterior3 / sum(posterior3) # standardize
#par(mfrow=c(1,3))
plot( posterior1 ~ p_grid , type="l" )
plot( posterior2 ~ p_grid , type="l" )
plot( posterior3 ~ p_grid , type="l" )
#par(mfrow=c(1,1))
```
  
  - 2M2: Use the following code:
```{r 2M2, echo=TRUE, message=FALSE, warning=FALSE}
p_grid <- seq( from=0 , to=1 , length.out=100 )
prior <- ifelse( p_grid < 0.5 , 0 , 1 ) # new prior
likelihood1 <- dbinom( 3 , size=3 , prob=p_grid ) # note this is for (1)
likelihood2 <- dbinom( 3 , size=4 , prob=p_grid ) # note this is for (2)
likelihood3 <- dbinom( 5 , size=7 , prob=p_grid ) # note this is for (3)
posterior1 <- likelihood1 * prior
posterior1 <- posterior1 / sum(posterior1) # standardize
posterior2 <- likelihood2 * prior
posterior2 <- posterior2 / sum(posterior2) # standardize
posterior3 <- likelihood3 * prior
posterior3 <- posterior3 / sum(posterior3) # standardize
#par(mfrow=c(1,3))
plot( posterior1 ~ p_grid , type="l" )
plot( posterior2 ~ p_grid , type="l" )
plot( posterior3 ~ p_grid , type="l" )
#par(mfrow=c(1,1))
```
  

  - 2M3: Expand Bayes' theorem
  
```{r 2M3, echo=TRUE, message=FALSE, warning=FALSE}
PLgE <- .3 # Prob of land given Earth:  70% water so 30% land
PLgM <- 1 # Prob of land given Mars:  all land hence no water

## equal priors
PE <- .5 # priors for Earth
PM <- .5 # priors for Mars

PEgL <- (PLgE * PE)/((PLgE*PE)+(PLgM*PM))
PEgL
```
  - 2M4: Do this problem at home

```{r 2M4, echo=TRUE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"

card.df <- data.frame(cards=1:3,nWaysB1=c(2,1,0),nWaysB2=c(2,0,0))

card.df

# add the second column
B1ways <- sum(card.df$nWaysB1)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```
  
  - 2M5: Same as before....
  
```{r 2M5, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","B")
c3 <- c("B","W")
c4 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:4,nWaysB1=c(2,2,1,0),nWaysB2=c(2,2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```

  - 2M6: Expand the counting table to include priors
```{r 2M6, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:3,priors=1:3,nWaysB1=c(2,1,0),nWaysB1Xpriors=c(2,2,0),nWaysB2=c(2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1Xpriors)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```
  - 2M7: Refer to the globe tossing counting table
```{r 2M7, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:3,priors=1:3,nWaysB1=c(2,1,0),nWaysB1Xpriors=c(2,2,0),nWaysB2=c(2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1Xpriors)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```  

### Hard ones:

  - 2H1: ballpark before you try to solve.  Here is my solution.  Begin with the what you are trying to solve for here:
  
  $$P(t2 | t1) = \frac{P(t1,t2)}{P(twins)}$$
where...

$$P(twins) = .5(.1) + .5(.2) = .15$$
Now, the probability from each species having two sets of twins requires multiplication and addition.

$$P(t1, t2) = .5(.1 * .1) + .5(.2 * .2) = 0.025$$
why?  Each set of twins is independent.  Thus, the probability of two twins back to back is merely the product of those two probabilities unconditional on anything else.

Now we can solve the problem because we have everything we need.

$$P(t2 | t1) = \frac{0.025}{0.15} \approx 0.17$$

  - 2H2:  Now compute P for the following conditional probability:
  
$$P(A | t1) = \frac{P(t1 | A) P(A)}{P(twins)}$$

We were given two bits already and computed one in the previous problem:

$$P(t1|A) = 0.1$$

$$P(A) = 0.5$$
recall that

$$P(twins) = 0.15$$
Now the problem is just a plug and chug:

$$P(A|t1) = \frac{P(t1 | A) P(A)}{P(twins)} = \frac{(0.1)(0.5)}{0.15} = \frac{1}{3}$$

  - 2H3: now we need to solve for a new probability but one that can be solved for in two ways.  The first way is to use all the available information and compute what is asked for in the question.  That probability is:
  
$$P(A | t1,!t2) = \frac{P(t1,!t2|A)P(A)}{P(t1,!t2)}$$

where $!t2$ represents the negation of twins at the second birth or rather a single birth.  I chose to be explicit with my notation but you could just as easily note the twins as with a T and the non-twins with an S.  

The denominator in the formula above is a critical one to remember because it is a definition that I keep emphasizing - the law of total probability.  We need to have the entire set included.  Thus....

$$P(t1,!t2) = P(t1,!t2 | A)P(A) + P(t1,!t2 | B)P(B)$$
You will see right away that the numerator is the first part and the alternative is the second part in the addition formula above.  

Now, let's get to solving some of these parts.

$$P(t1,!t2 | A) = (0.1)(1-0.1) = (0.1)(0.9) = 0.09$$
the second bit is similar but only pertains to species B:

$$P(t1,!t2 | B) = (0.2)(1-0.2) = (0.2)(0.8) = 0.16$$
Recall that the priors for each species remains the same.  As such:

$$P(A) = P(B) = 0.5$$

and...

$$P(t1,!t2) = P(t1,!t2 | A)P(A) + P(t1,!t2 | B)P(B) = (0.09)(0.5) + (0.16)(0.5)$$
$$P(t1,!t2) = 0.125$$

We now use this as our denominator in our original equation:

$$P(A | t1,!t2) = \frac{(0.09)(0.5)}{0.125} = 0.36$$


# Chapter 3

Consider the examples I used above but here I tell a story about natural frequencies relevant to our problem of estimation.  Here goes...

1. 1700 universities with psychology programs and less than half conduct research ($\approx 800$)
2. Of the 800 programs that conduct research, roughly 10 faculty per department (on average) are active researchers; thus $N_{researchers} \approx 8000$.
3. 5% of active researchers contribute more than 80% of the research; thus, we can safely say that our real $N_{researchers}$ is about 400.
4. Of those 400 researchers, each contribute about 5 empirical, peer-reviewed papers each year; thus, these researchers produce 2000 articles each year.
5. Retraction watch lists 225 articles from 1999 to the present (22 years or roughly 10 articles each year)
6. Assume for every article found with problems, 9 others exist; thus, 100 articles each year should be a good estimate of fraud.  $N_{fraud}=100$
7. Those who commit scientific fraud are likely to do so repeatedly and for all their publications.  Thus, not all events are independent and we get an estimate of researchers who commit fraud to be $N_{fraudsters} = N_{researchers}/N_{fraud} = 4$

Use these numbers to tell a story about fraudsters.  Use only the frequencies.

## Assigned Problems:

### Easy Preamble

The Easy problems use the samples from the posterior distribution for the globe tossing example. This code will give you a specific set of samples, so that you can check your answers exactly.

#### 3E1:  How much posterior probability lies below p = 0.2?

```{r 3E1}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
sum(samples < 0.2)/1e4
```

#### 3E2: How much posterior probability lies above p=0.8?

```{r 3E2}
sum(samples > 0.8)/ 1e4
```

#### 3E3:  How much posterior probability lies between .2 and .8?

```{r 3E3}
sum( samples > 0.2 & samples < 0.8 ) / 1e4
```

#### 3E4: 20% of the posterior probability lies below which probability value?

```{r 3E4}
quantile( samples , 0.2 ) ## get the prob first, round to 2 digits
sum( samples < 0.52) / 1e4 ## calculate with the new rounded value
```

#### 3E5: 20% of the posterior probability lies above which value of p?

```{r 3E5}
quantile( samples , 0.8 ) # reverse logic of above, round to 2 digits
sum( samples > 0.76) / 1e4 ## calculate with the new rounded value
```


#### 3E6: Which values of p contain the narrowest interval equal to 66% of the posterior probability?

```{r 3E6}
HPDI(samples, prob=0.66)
```

#### 3E7: Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r 3E7}
PI(samples,prob=0.66)
```
### Medium Difficulty Problems

#### 3M1: Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r 3M1}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1,1000)
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot(posterior ~ p_grid, type="l")
```

#### 3M2: Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.

```{r 3M2}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
HPDI(samples, prob=0.9)
```

#### 3M3: Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r 3M3}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
water <- rbinom(1e4, size=15, prob=samples)
sum(water==8)/1e4
simplehist(water)
```

#### 3M4: Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

```{r 3M4}
water <- rbinom(1e4, size=9, prob=samples)
sum(water == 6) / 1e4
```

#### 3M5: Start over at 3M1, but now use a prior that is zero below 0.5 and a constant above 0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value 0.7.

```{r 3M5}
p_grid <- seq(0,1,length.out=1000)
prior <- ifelse(p_grid < .5, 0, 1) # new prior
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
plot(posterior ~ p_grid, type="l")
HPDI (samples, prob=0.9)
```

#### 3M6: Suppose you want to estimate the Earth’s proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this?

```{r 3M6}
water <- rbinom(1e4, size=15, prob=samples)
simplehist(water)
```

### Hard questions

The Hard problems here all use the data below. These data indicate the gender (male=1, fe-male=0) of officially reported first and second born children in 100 two-child families.

```{r 3H_data}
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,1,0,1,1,1,0,1,1,1,1)
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,0,0,0,1,1,1,0,0,0,0)

## make sure you load the following:
library(rethinking)
data(homeworkch3) # alternative to reading in two vectors above
```

#### 3H1: Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

```{r 3H1}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1,1000) ## note, you can use length(p_grid) as well
boys <- sum(birth1) + sum(birth2)
likelihood <- dbinom(boys, size=200, prob= p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot(posterior ~ p_grid, type="l")
abline(v=0.5, lty=2)
p_grid[which.max(posterior)]
```

#### 3H2: Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals.

```{r 3H2}
p.samples <- sample(p_grid, size=10000, replace = TRUE, prob = posterior)
HPDI(p.samples, prob = 0.5)
HPDI(p.samples, prob = 0.89)
HPDI(p.samples, prob = 0.97)
```

#### 3H3: Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to visualize the simulations, but the dens command (part of the rethinking package) is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

```{r 3H3}
p.samples <- sample(p_grid, size=10000, replace=TRUE, prob=posterior)
bsim <- rbinom(10000, size=200, prob=p.samples)
dens(bsim, adj=0.1)
abline(v=sum(birth1)+sum(birth2), col="red")
```

#### 3H4: Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1. How does the model look in this light?

```{r 3H4}
b1sim <- rbinom(10000, size=100, prob=p.samples)
dens(b1sim, adj=0.1)
abline(v=sum(birth1), col="red")
```

#### 3H5: The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births, 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses what is going on in these data?

```{r 3H5}
b01 <- birth2[birth1==0]
b01sim <- rbinom(10000, size=length(b01), prob=p.samples)
dens(b01sim, adj=0.1)
abline(v=sum(b01), col="red")
```


# Chapter 4

Key points:

1. Differentiate between observable (data) and unobservable (parameters) variables.  Consider the simple MRC model:

$$y = bx + b_0 + \epsilon \tag{full MRC model}$$

$$y = bx + b_0 \tag{deterministic MRC model}$$
$...$that we rewrite in more explicit form:

$$y_i = \beta (x_i - \bar{x})  + \beta_0$$
  $...$where$...$
    $y_i$ and $x_i$ are both observed and $\beta$ and $\beta_0$ are parameters.  

2. Define each variable in terms of other variables OR in terms of a probability distribution.  (see page 96 for all the definitions)

$$y_i \sim \mathcal{N}(\mu_i, \sigma) \tag{The Likelihood}$$
rewritten in R as: y $\sim$ dnorm(mu, sigma)

$$\mu_i = \beta (x_i-\bar{x}) + \beta_0 \tag{the GLM}$$
rewritten in R as: mu <- b*x + b0

$$\beta \sim \mathcal{N}(0,4) \tag{beta prior}$$
rewritten in R as: b $\sim$ dnorm(0,4)

$$\beta_0 \sim \mathcal{N}(200,20) \tag{intercept prior}$$
rewritten in R as: b0 $\sim$ dnorm(200,20)

$$\sigma \sim Uniform(0,10) \tag{sigma prior}$$
and finally, rewritten in R as:  sigma $\sim$ dunif(0,10)

See code chunk 4.42 for comparison.

3. The combination from above defines a *joint generative model* - one that can be used to (a) simulate hypothetical observations and (b) analyze real ones.  This last point is where Bayesian methods take a strong departure from frequentist models.  Note the following:

* extract.samples() is a function that helps us create data from the results.
* precis() is a function that helps us summarize data.frames (and samples)
* mvrnorm() is a function that allows us to sample conditionally on multiple variables/parameters

The rest of the chapter builds upon these ideas while also introducing the reader to more R details.  If you are new to R, please read these sections carefully.  Here are some highlights and points of interest....

* square brackets $[x]$ or $[x,y]$ are key to subsetting vectors and data.frames, respectively.
* list and alist are two different but important storage containers.  Know them.
* rnorm and rlnorm are functions that create random normal and log-normal values.
* plot your priors and posteriors! Learn base graphics (plot) AND ggplot
* be careful when reading Richard's explanation of the intercept (esp p 115)
* polynomial and spline results are tricky in psychological science.  Let's discuss!
* the code in 4.76 requires you to know one teeny weeny aspect of linear algebra.  I'll elaborate in class.

## Assigned Problems

### Easy Problems

#### 4E1:  In the model definition below, which line is the likelihood?
Likelihood is the conditional statement on $y_i$

#### 4E2: In the model definition just above, how many parameters are in the posterior distribution?
Two - $\mu$ and $\sigma$.

#### 4E3: Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.

Not essential but useful if you want to keep track of the formulae he uses throughout the book.  Here goes...

From Section 4.3.2 (Overthinking: Model definition to Bayes' theorem again.) in the book, take the formula and replace his bits with the bits provided in the example.  Why do you need to know this?  Dunno.  You will never remember it nor will you every engage in integral calculus.  


#### 4E4: In the model definition below, which line is the linear model?

The second line.


#### 4E5:  In the model definition just above, how many parameters are in the posterior distribution?

Three:  $\alpha$, $\beta$, and $\sigma$.

#### 4M1: For the model definition below, simulate observed y values from the prior (not the posterior).

```{r 4M1}
mu_prior <- rnorm(1e4, 0, 10)
sigma_prior <- rexp(1e4, 1)
h_sim <- rnorm(1e4, mu_prior, sigma_prior)
dens(h_sim)
```


#### 4M2: Translate the model just above into a quap formula.

```{r 4M2}
f <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1)
)
```

#### 4M3: Translate the quap model formula below into a mathematical model definition.

$$y_i \sim Normal(\mu, \sigma)$$
$$\mu = \alpha + \beta x_i $$
$$\alpha \sim Normal(0,10) $$
$$\beta \sim Normal(0,1)$$
$$\sigma \sim Exponential(1)$$

#### 4M4: A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

Assume the following model:


$$h_{ij} \sim Normal(\mu_{ij}, \sigma)$$
$$\mu_{ij} = \alpha + \beta(y_i - \bar{y})$$
$$\alpha \sim Normal(68, 10)$$
$$\beta \sim Normal(0,10)$$
$$\sigma \sim Exponential(1)$$
and in R to simulate:

```{r 4M4}
n <- 100
a <- rnorm(n, 68, 10)
b <- rnorm(n, 0, 10)
s <- rexp(n, 1)
y <- 1:3
ybar <- mean(y)
h <- matrix(NA, nrow=n, ncol=3)
for (i in 1:n){
  for (j in 1:3){
    h[i,j] <- rnorm(1, a[i] + b[i]*(y[j] - ybar), s[i])
  }
}
plot(NULL, xlim=c(1,3), ylim=range(h), xlab="Year", ylab="Height (inches)")

for(i in 1:n){
  lines(1:3, h[i,])
}
```

Whoops!  Some students are getting shorter.  

#### 4M5: Now suppose I remind you that every student got taller each year. Does this information lead you to change your choice of priors? How?

Yes, of course.  I messed up with my previous formulation and allowed $\beta$ to be negative.  Whoopsi!  OK, a new method...

$$h_{ij} \sim Normal(\mu_{ij}, \sigma)$$
$$\mu_{ij} = \alpha + \beta(y_i - \bar{y})$$
$$\alpha \sim Normal(68, 10)$$
$$\beta \sim Log-Normal(1,0.5)$$
$$\sigma \sim Exponential(1)$$



```{r 4M5}
n <- 100
a <- rnorm(n, 68, 10)
b <- rlnorm(n, 1, 0.5)
s <- rexp(n, 1)
y <- 1:3
ybar <- mean(y)
h <- matrix(NA, nrow=n, ncol=3)
for (i in 1:n){
  for (j in 1:3){
    h[i,j] <- rnorm(1, a[i] + b[i]*(y[j] - ybar), s[i])
  }
}
plot(NULL, xlim=c(1,3), ylim=range(h), xlab="Year", ylab="Height (inches)")

for(i in 1:n){
  lines(1:3, h[i,])
}


```

#### 4M6: Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm (25 inches). How does this lead you to revise your priors?

$$\sigma \sim Uniform(0,5)$$

because...

$$ \sigma^2 = 25$$
but...

$$\sigma = \sqrt{25} = 5$$

#### 4M7: Refit model m4.3 from the chapter, but omit the mean weight xbar this time. Compare the new model’s posterior to that of the original model. In particular, look at the covariance among the parameters. What is different? Then compare the posterior predictions of both models.

```{r 4M7}
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18,]
m4.3b <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*weight,
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ), data = d2
)
precis(m4.3b)
```

Compare with the original model of m4.3.  Notice the differences in intercepts?

```{r}
round(vcov(m4.3b),2)
```
```{r}
round(cov2cor(vcov(m4.3b)),2)
```


#### 4M8: In the chapter, we used 15 knots with the cherry blossom spline. Increase the number of knots and observe what happens to the resulting spline. Then adjust also the width of the prior on the weights—change the standard deviation of the prior and watch what happens. What do you think the combination of knot number and the prior on the weights controls?

```{r 4M8}

```


#### 4H1: The weights listed below were recorded in the !Kung census, but heights were not recorded for these individuals. Provide predicted heights and 89% intervals for each of these individuals. That is, fill in the table below, using model-based predictions.

```{r 4H1}

```


#### 4H2: Select out all the rows in the Howell1 data with ages below 18 years of age. If you do it right, you should end up with a new data frame with 192 rows in it.
(a)  Fit a linear regression to these data, using quap. Present and interpret the estimates. For every 10 units of increase in weight, how much taller does the model predict a child gets?
(b)  Plot the raw data, with height on the vertical axis and weight on the horizontal axis. Superimpose the MAP regression line and 89% interval for the mean. Also superimpose the 89% interval for predicted heights.
(c)  What aspects of the model fit concern you? Describe the kinds of assumptions you would change, if any, to improve the model. You don’t have to write any new code. Just explain what the model appears to be doing a bad job of, and what you hypothesize would be a better model.


#### 4H3: Suppose a colleague of yours, who works on allometry, glances at the practice problems just above. Your colleague exclaims, “That’s silly. Everyone knows that it’s only the logarithm of body weight that scales with height!” Let’s take your colleague’s advice and see what happens.

(a)  Model the relationship between height (cm) and the natural logarithm of weight (log-kg). Use the entire Howell1 data frame, all 544 rows, adults and non-adults. Can you interpret the resulting estimates?

(b)  Begin with this plot: plot( height ~ weight , data=Howell1 ). Then use samples from the quadratic approximate posterior of the model in (a) to superimpose on the plot: (1) the predicted mean height as a function of weight, (2) the 97% interval for the mean, and (3) the 97% interval for predicted heights.
#### 4H4: Plot the prior predictive distribution for the parabolic polynomial regression model in the chapter. You can modify the code that plots the linear regression prior predictive distribution. Can you modify the prior distributions of α, β1, and β2 so that the prior predictions stay within the biologically reasonable outcome space? That is to say: Do not try to fit the data by hand. But do try to keep the curves consistent with what you know about height and weight, before seeing these exact data.
#### 4H5: Return to data(cherry_blossoms) and model the association between blossom date (doy) and March temperature (temp). Note that there are many missing values in both variables. You may consider a linear model, a polynomial, or a spline on temperature. How well does temperature trend predict the blossom trend?
#### 4H6:  Simulate the prior predictive distribution for the cherry blossom spline in the chapter. Adjust the prior on the weights and observe what happens. What do you think the prior on the weights is doing?
#### 4H7: TYPO - no 7th Hard problem
#### 4H8: The cherry blossom spline in the chapter used an intercept α, but technically it doesn’t require one. The first basis functions could substitute for the intercept. Try refitting the cherry blossom spline without the intercept. What else about the model do you need to change to make this work?

# Chapter 5

$$A \perp \!\!\! \perp B $$
Some key terms to consider:
* spurious relationships
* confounded relationships
* multicollinearity
* 
