---
title: "Patrick's PSYC 757 Class Notes"
date: "`r paste('Last edits on:',date(), sep=' ')`"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document: default
  html_notebook: default
---

```{r preamble, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	background = "#ADD8E6",
	cache = TRUE,
	tidy = 'styler',
	class.source = "bg-warning",
	class.output = "bg-success",
	fig_caption = TRUE,
	digits = 3
)


```

```{r libraries, include=FALSE}
using<-function(...) {
    libs<-unlist(list(...))
    req<-unlist(lapply(libs,require,character.only=TRUE))
    need<-libs[req==FALSE]
    if(length(need)>0){ 
        install.packages(need)
        lapply(need,require,character.only=TRUE)
    }
}
using("tidyverse","ggplot2","psych","corrplot","DT","formatR","ggExtra","car","rethinking","rstan","rvest","dagitty")

round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))

  df[,nums] <- round(df[,nums], digits = digits)

  (df)
}

```

# Chapter 1

Prepare yourself to learn statistics from a different angle.  Also, get ready to learn some skills (and tools).  Those include R, Rstudio, stan, and for the daring, tidyverse.

## Getting started

1. Install R - get the latest version for your platform
2. Install RStudio - get the **preview** version instead of the standard download.
3. Install rstan - see [mc-stan.org](https://mc-stan.org) for details

## Getting prepared each week

1. Read the assigned book chapter.  I assume all graduate students read so please do so.
2. Watch the videos.  If you find an additional loop through the readings helps, I strongly encourage you to watch them.  Richard does a fine job providing further details from his book.
3. Come prepared to ask questions on Tuesdays.
4. Come prepared to do some homework together - Tuesdays (occasionally) and Thursdays (always).
5. Stay ahead of the workflow.  
6. Check for additional references including the [R code from the book](http://xcelab.net/rmpubs/sr2/code.txt) and the efforts to port that code to more modern tools.  Please consult [Richard's webpage](https://xcelab.net/rm/statistical-rethinking/) for more links.

# Chapter 2

Probability theory requires some work.  We will work on a problem together and then solve the chapter problems together on Thursday.  Here are the problems I want us to work on together during the Tuesday lecture period:

## Some fundamentals

[Look here](https://www.statlect.com/) for some useful information during the course.

1. Definitions:
  - Probability refers to an event (i.e., something happens): $P(A)$
  - Probability is a number (but it can be qualified later as a qualitative feel) and ranges between 0 and 1:  $0 \le P(A) \le 1$
  - Rules define how we compute and evaluate probabilities
2. Probabilities are ratios of counts...
  - An event occurring count (Numerator)
  - Total count of observations (Denominator)
  - Thus: $p = Event_{count} / (Event_{count} + Non-Event_{count})$
3. Arithmetic on probabilities are easy when...
  - Events are independent
  - Total count of observations are readily available
4. Rules of probability (you will learn these implicitly over time)
  - Addition Rule: $$P(A or B) = P(A \cup B) = P(A) + P(B) - P(A \cap B)$$
  - Complementary Rule: $$P(A) + P(A^{\prime}) = 1$$
  - Conditional Rule: $$P(B, given A) = P(B | A)=P(A \cap B) / P(A)$$
  - Multiplication Rule: $$P(A and B) = P(A \cap B) = P(A) \cdot P(B|A) $$

## An Example

Let's attack psychological science since we are all in the game now.  We appear to be in a crisis where most if not all our published results are either being retracted or questioned based upon glaring failures to replicate.  If a study fails to replicate, should we be surprised?  And, would this surprise lead us to feel as if we are in crisis?  Here are some numbers:

1. Assumptions
  - we are all frequentists and employ some form of NHST (e.g., ANOVA, MRC, etc.).
  - we use a simple probability of 5% error to define "significance" (i.e., $p \le 0.05$).
  - Expected Base Rate (priors condition):  we only venture where it is safe to venture ($P(H_{ours}) \ge P(H_{alts})$)
  - Evidence Discovery Rate:  we only test what is readily available to us ($P(D_{ours}) \ge .5$)
  - True Positive Rate:  we measure with almost perfect validity and precision (a point I shall elaborate upon) but for now, we specify it as a probability ($P(Reject H_0 | H_0 = F) = .99$, aka statistical power or, for our question though...$P(D_{ours} = T | H_{ours}=T) = .99$)
  - False Positive Rate:  we have a very low false positive rate ($P(D_{ours} = T| H_{ours} = F) \le .01$)
  - every study is independent of every other study - including our own
  - subjects are replicates of one another so sample-to-sample variance is zero
  - Base Rate for $H_{ours}$: there are 100 hypotheses available for the data in question and ours is only one of the 100 including the null ($P(H_{ours}) \approx .02$ or double for each $P(H_{alt}$)
2. Computations
  - Use Bayes Theorem with limited information
  $$P(H_{ours}|D) = \frac{P(D_{ours}|H_{ours}) \cdot P(H_{ours}))}{P(D)}$$
  $$P(H_{ours}|D) = \frac{.99 * .02}{.5} = 0.0396 \approx 4\%$$
  - Alternative with full information
  $$P(H_{ours}|D) = \frac{P(D_{ours}|H_{ours}=T) \cdot P(H_{ours}))}{P(D_{ours}|H_{ours}) \cdot P(H_{ours}) + P(D_{ours}| H_{ours} = F) \cdot (1 - P(H_{ours}))}$$
  $$P(H_{ours}|D) = \frac{.99 * .02}{(.99 * .02)+(.01 * .98)} = 0.669 \approx 67\%$$
  - Why the discrepency?
  Consider Richard's example.  He used $P(D_{ours} | H_{ours} = T) = 0.95$, $P(D_{ours} | H_{ours} = F) = 0.05$, $P(H_{ours}) = 0.01$, gives us:
  $$P(H_{ours}|D) = \frac{0.95 * 0.01}{(0.95 * 0.01)+(0.05 * 0.99)} = 0.161 \approx 16\%$$
  - Use your own numbers and see how well you can make the replication crisis seem worse or better.

## Assigned Problems:

### All Easy ones (hints below)
  - 2E1: Notation and Definitions
  - 2E2: Notation
  - 2E3: Notation and Definitions
  - 2E4: Definitions and Intentions of Probability

### Medium ones:
  - 2M1: Use the following code (revealed after class) to produce the following:
```{r 2M1, echo=TRUE, message=FALSE, warning=FALSE}
p_grid <- seq( from=0 , to=1 , length.out=100 )
# likelihood of 3 water in 3 tosses
likelihood1 <- dbinom( 3 , size=3 , prob=p_grid ) # note this is for (1)
likelihood2 <- dbinom( 3 , size=4 , prob=p_grid ) # note this is for (2)
likelihood3 <- dbinom( 5 , size=7 , prob=p_grid ) # note this is for (3)
prior <- rep(1,100) # uniform prior
posterior1 <- likelihood1 * prior
posterior1 <- posterior1 / sum(posterior1) # standardize
posterior2 <- likelihood2 * prior
posterior2 <- posterior2 / sum(posterior2) # standardize
posterior3 <- likelihood3 * prior
posterior3 <- posterior3 / sum(posterior3) # standardize
#par(mfrow=c(1,3))
plot( posterior1 ~ p_grid , type="l" )
plot( posterior2 ~ p_grid , type="l" )
plot( posterior3 ~ p_grid , type="l" )
#par(mfrow=c(1,1))
```
  
  - 2M2: Use the following code:
```{r 2M2, echo=TRUE, message=FALSE, warning=FALSE}
p_grid <- seq( from=0 , to=1 , length.out=100 )
prior <- ifelse( p_grid < 0.5 , 0 , 1 ) # new prior
likelihood1 <- dbinom( 3 , size=3 , prob=p_grid ) # note this is for (1)
likelihood2 <- dbinom( 3 , size=4 , prob=p_grid ) # note this is for (2)
likelihood3 <- dbinom( 5 , size=7 , prob=p_grid ) # note this is for (3)
posterior1 <- likelihood1 * prior
posterior1 <- posterior1 / sum(posterior1) # standardize
posterior2 <- likelihood2 * prior
posterior2 <- posterior2 / sum(posterior2) # standardize
posterior3 <- likelihood3 * prior
posterior3 <- posterior3 / sum(posterior3) # standardize
#par(mfrow=c(1,3))
plot( posterior1 ~ p_grid , type="l" )
plot( posterior2 ~ p_grid , type="l" )
plot( posterior3 ~ p_grid , type="l" )
#par(mfrow=c(1,1))
```
  

  - 2M3: Expand Bayes' theorem
  
```{r 2M3, echo=TRUE, message=FALSE, warning=FALSE}
PLgE <- .3 # Prob of land given Earth:  70% water so 30% land
PLgM <- 1 # Prob of land given Mars:  all land hence no water

## equal priors
PE <- .5 # priors for Earth
PM <- .5 # priors for Mars

PEgL <- (PLgE * PE)/((PLgE*PE)+(PLgM*PM))
PEgL
```
  - 2M4: Do this problem at home

```{r 2M4, echo=TRUE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"

card.df <- data.frame(cards=1:3,nWaysB1=c(2,1,0),nWaysB2=c(2,0,0))

card.df

# add the second column
B1ways <- sum(card.df$nWaysB1)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```
  
  - 2M5: Same as before....
  
```{r 2M5, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","B")
c3 <- c("B","W")
c4 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:4,nWaysB1=c(2,2,1,0),nWaysB2=c(2,2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```

  - 2M6: Expand the counting table to include priors
```{r 2M6, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:3,priors=1:3,nWaysB1=c(2,1,0),nWaysB1Xpriors=c(2,2,0),nWaysB2=c(2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1Xpriors)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```
  - 2M7: Refer to the globe tossing counting table
```{r 2M7, echo=FALSE, message=FALSE, warning=FALSE}
c1 <- c("B","B")
c2 <- c("B","W")
c3 <- c("W","W")

# 3 cards produce 3 ways for the side shown to be "B"
card.df <- data.frame(cards=1:3,priors=1:3,nWaysB1=c(2,1,0),nWaysB1Xpriors=c(2,2,0),nWaysB2=c(2,0,0))
card.df

# add the second column
B1ways <- sum(card.df$nWaysB1Xpriors)
B2ways <- sum(card.df$nWaysB2)
B2ways/B1ways
```  

### Hard ones:

  - 2H1: ballpark before you try to solve.  Here is my solution.  Begin with the what you are trying to solve for here:
  
  $$P(t2 | t1) = \frac{P(t1,t2)}{P(twins)}$$
where...

$$P(twins) = .5(.1) + .5(.2) = .15$$
Now, the probability from each species having two sets of twins requires multiplication and addition.

$$P(t1, t2) = .5(.1 * .1) + .5(.2 * .2) = 0.025$$
why?  Each set of twins is independent.  Thus, the probability of two twins back to back is merely the product of those two probabilities unconditional on anything else.

Now we can solve the problem because we have everything we need.

$$P(t2 | t1) = \frac{0.025}{0.15} \approx 0.17$$

  - 2H2:  Now compute P for the following conditional probability:
  
$$P(A | t1) = \frac{P(t1 | A) P(A)}{P(twins)}$$

We were given two bits already and computed one in the previous problem:

$$P(t1|A) = 0.1$$

$$P(A) = 0.5$$
recall that

$$P(twins) = 0.15$$
Now the problem is just a plug and chug:

$$P(A|t1) = \frac{P(t1 | A) P(A)}{P(twins)} = \frac{(0.1)(0.5)}{0.15} = \frac{1}{3}$$

  - 2H3: now we need to solve for a new probability but one that can be solved for in two ways.  The first way is to use all the available information and compute what is asked for in the question.  That probability is:
  
$$P(A | t1,!t2) = \frac{P(t1,!t2|A)P(A)}{P(t1,!t2)}$$

where $!t2$ represents the negation of twins at the second birth or rather a single birth.  I chose to be explicit with my notation but you could just as easily note the twins as with a T and the non-twins with an S.  

The denominator in the formula above is a critical one to remember because it is a definition that I keep emphasizing - the law of total probability.  We need to have the entire set included.  Thus....

$$P(t1,!t2) = P(t1,!t2 | A)P(A) + P(t1,!t2 | B)P(B)$$
You will see right away that the numerator is the first part and the alternative is the second part in the addition formula above.  

Now, let's get to solving some of these parts.

$$P(t1,!t2 | A) = (0.1)(1-0.1) = (0.1)(0.9) = 0.09$$
the second bit is similar but only pertains to species B:

$$P(t1,!t2 | B) = (0.2)(1-0.2) = (0.2)(0.8) = 0.16$$
Recall that the priors for each species remains the same.  As such:

$$P(A) = P(B) = 0.5$$

and...

$$P(t1,!t2) = P(t1,!t2 | A)P(A) + P(t1,!t2 | B)P(B) = (0.09)(0.5) + (0.16)(0.5)$$
$$P(t1,!t2) = 0.125$$

We now use this as our denominator in our original equation:

$$P(A | t1,!t2) = \frac{(0.09)(0.5)}{0.125} = 0.36$$


# Chapter 3

Consider the examples I used above but here I tell a story about natural frequencies relevant to our problem of estimation.  Here goes...

1. 1700 universities with psychology programs and less than half conduct research ($\approx 800$)
2. Of the 800 programs that conduct research, roughly 10 faculty per department (on average) are active researchers; thus $N_{researchers} \approx 8000$.
3. 5% of active researchers contribute more than 80% of the research; thus, we can safely say that our real $N_{researchers}$ is about 400.
4. Of those 400 researchers, each contribute about 5 empirical, peer-reviewed papers each year; thus, these researchers produce 2000 articles each year.
5. Retraction watch lists 225 articles from 1999 to the present (22 years or roughly 10 articles each year)
6. Assume for every article found with problems, 9 others exist; thus, 100 articles each year should be a good estimate of fraud.  $N_{fraud}=100$
7. Those who commit scientific fraud are likely to do so repeatedly and for all their publications.  Thus, not all events are independent and we get an estimate of researchers who commit fraud to be $N_{fraudsters} = N_{researchers}/N_{fraud} = 4$

Use these numbers to tell a story about fraudsters.  Use only the frequencies.

## Assigned Problems:

### Easy Preamble

The Easy problems use the samples from the posterior distribution for the globe tossing example. This code will give you a specific set of samples, so that you can check your answers exactly.

#### 3E1:  How much posterior probability lies below p = 0.2?

```{r 3E1}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
sum(samples < 0.2)/1e4
```

#### 3E2: How much posterior probability lies above p=0.8?

```{r 3E2}
sum(samples > 0.8)/ 1e4
```

#### 3E3:  How much posterior probability lies between .2 and .8?

```{r 3E3}
sum( samples > 0.2 & samples < 0.8 ) / 1e4
```

#### 3E4: 20% of the posterior probability lies below which probability value?

```{r 3E4}
quantile( samples , 0.2 ) ## get the prob first, round to 2 digits
sum( samples < 0.52) / 1e4 ## calculate with the new rounded value
```

#### 3E5: 20% of the posterior probability lies above which value of p?

```{r 3E5}
quantile( samples , 0.8 ) # reverse logic of above, round to 2 digits
sum( samples > 0.76) / 1e4 ## calculate with the new rounded value
```


#### 3E6: Which values of p contain the narrowest interval equal to 66% of the posterior probability?

```{r 3E6}
HPDI(samples, prob=0.66)
```

#### 3E7: Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r 3E7}
PI(samples,prob=0.66)
```
### Medium Difficulty Problems

#### 3M1: Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r 3M1}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1,1000)
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot(posterior ~ p_grid, type="l")
```

#### 3M2: Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.

```{r 3M2}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
HPDI(samples, prob=0.9)
```

#### 3M3: Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r 3M3}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
water <- rbinom(1e4, size=15, prob=samples)
sum(water==8)/1e4
simplehist(water)
```

#### 3M4: Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

```{r 3M4}
water <- rbinom(1e4, size=9, prob=samples)
sum(water == 6) / 1e4
```

#### 3M5: Start over at 3M1, but now use a prior that is zero below 0.5 and a constant above 0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value 0.7.

```{r 3M5}
p_grid <- seq(0,1,length.out=1000)
prior <- ifelse(p_grid < .5, 0, 1) # new prior
likelihood <- dbinom(8, size=15, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
plot(posterior ~ p_grid, type="l")
HPDI (samples, prob=0.9)
```

#### 3M6: Suppose you want to estimate the Earth’s proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this?

```{r 3M6}
water <- rbinom(1e4, size=15, prob=samples)
simplehist(water)
```

### Hard questions

The Hard problems here all use the data below. These data indicate the gender (male=1, fe-male=0) of officially reported first and second born children in 100 two-child families.

```{r 3H_data}
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,1,0,1,1,1,0,1,1,1,1)
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,0,0,0,1,1,1,0,0,0,0)

## make sure you load the following:
library(rethinking)
data(homeworkch3) # alternative to reading in two vectors above
```

#### 3H1: Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

```{r 3H1}
p_grid <- seq(0,1,length.out=1000)
prior <- rep(1,1000) ## note, you can use length(p_grid) as well
boys <- sum(birth1) + sum(birth2)
likelihood <- dbinom(boys, size=200, prob= p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
plot(posterior ~ p_grid, type="l")
abline(v=0.5, lty=2)
p_grid[which.max(posterior)]
```

#### 3H2: Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals.

```{r 3H2}
p.samples <- sample(p_grid, size=10000, replace = TRUE, prob = posterior)
HPDI(p.samples, prob = 0.5)
HPDI(p.samples, prob = 0.89)
HPDI(p.samples, prob = 0.97)
```

#### 3H3: Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to visualize the simulations, but the dens command (part of the rethinking package) is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

```{r 3H3}
p.samples <- sample(p_grid, size=10000, replace=TRUE, prob=posterior)
bsim <- rbinom(10000, size=200, prob=p.samples)
dens(bsim, adj=0.1)
abline(v=sum(birth1)+sum(birth2), col="red")
```

#### 3H4: Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1. How does the model look in this light?

```{r 3H4}
b1sim <- rbinom(10000, size=100, prob=p.samples)
dens(b1sim, adj=0.1)
abline(v=sum(birth1), col="red")
```

#### 3H5: The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births, 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses what is going on in these data?

```{r 3H5}
b01 <- birth2[birth1==0]
b01sim <- rbinom(10000, size=length(b01), prob=p.samples)
dens(b01sim, adj=0.1)
abline(v=sum(b01), col="red")
```


# Chapter 4

Key points:

1. Differentiate between observable (data) and unobservable (parameters) variables.  Consider the simple MRC model:

$$y = bx + b_0 + \epsilon \tag{full MRC model}$$

$$y = bx + b_0 \tag{deterministic MRC model}$$
$...$that we rewrite in more explicit form:

$$y_i = \beta (x_i - \bar{x})  + \beta_0$$
  $...$where$...$
    $y_i$ and $x_i$ are both observed and $\beta$ and $\beta_0$ are parameters.  

2. Define each variable in terms of other variables OR in terms of a probability distribution.  (see page 96 for all the definitions)

$$y_i \sim \mathcal{N}(\mu_i, \sigma) \tag{The Likelihood}$$
rewritten in R as: y $\sim$ dnorm(mu, sigma)

$$\mu_i = \beta (x_i-\bar{x}) + \beta_0 \tag{the GLM}$$
rewritten in R as: mu <- b*x + b0

$$\beta \sim \mathcal{N}(0,4) \tag{beta prior}$$
rewritten in R as: b $\sim$ dnorm(0,4)

$$\beta_0 \sim \mathcal{N}(200,20) \tag{intercept prior}$$
rewritten in R as: b0 $\sim$ dnorm(200,20)

$$\sigma \sim Uniform(0,10) \tag{sigma prior}$$
and finally, rewritten in R as:  sigma $\sim$ dunif(0,10)

See code chunk 4.42 for comparison.

3. The combination from above defines a *joint generative model* - one that can be used to (a) simulate hypothetical observations and (b) analyze real ones.  This last point is where Bayesian methods take a strong departure from frequentist models.  Note the following:

* extract.samples() is a function that helps us create data from the results.
* precis() is a function that helps us summarize data.frames (and samples)
* mvrnorm() is a function that allows us to sample conditionally on multiple variables/parameters

The rest of the chapter builds upon these ideas while also introducing the reader to more R details.  If you are new to R, please read these sections carefully.  Here are some highlights and points of interest....

* square brackets $[x]$ or $[x,y]$ are key to subsetting vectors and data.frames, respectively.
* list and alist are two different but important storage containers.  Know them.
* rnorm and rlnorm are functions that create random normal and log-normal values.
* plot your priors and posteriors! Learn base graphics (plot) AND ggplot
* be careful when reading Richard's explanation of the intercept (esp p 115)
* polynomial and spline results are tricky in psychological science.  Let's discuss!
* the code in 4.76 requires you to know one teeny weeny aspect of linear algebra.  I'll elaborate in class.

## Assigned Problems

### Easy Problems

#### 4E1:  In the model definition below, which line is the likelihood?
Likelihood is the conditional statement on $y_i$

#### 4E2: In the model definition just above, how many parameters are in the posterior distribution?
Two - $\mu$ and $\sigma$.

#### 4E3: Using the model definition above, write down the appropriate form of Bayes’ theorem that includes the proper likelihood and priors.

Not essential but useful if you want to keep track of the formulae he uses throughout the book.  Here goes...

From Section 4.3.2 (Overthinking: Model definition to Bayes' theorem again.) in the book, take the formula and replace his bits with the bits provided in the example.  Why do you need to know this?  Dunno.  You will never remember it nor will you every engage in integral calculus.  


#### 4E4: In the model definition below, which line is the linear model?

The second line.


#### 4E5:  In the model definition just above, how many parameters are in the posterior distribution?

Three:  $\alpha$, $\beta$, and $\sigma$.

### Medium Problems

#### 4M1: For the model definition below, simulate observed y values from the prior (not the posterior).

```{r 4M1, echo=TRUE}
library(rethinking)
mu_prior <- rnorm(1e4, 0, 10)
dens(mu_prior)
sigma_prior <- rexp(1e4, 1)
h_sim <- rnorm(1e4, mu_prior, sigma_prior)
dens(h_sim)
```


#### 4M2: Translate the model just above into a quap formula.

```{r 4M2}
f <- alist(
  y ~ dnorm(mu, sigma),
  mu ~ dnorm(0,10),
  sigma ~ dexp(1)
)
```

#### 4M3: Translate the quap model formula below into a mathematical model definition.

$$y_i \sim Normal(\mu, \sigma)$$
$$\mu = \alpha + \beta x_i $$
$$\alpha \sim Normal(0,10) $$
$$\beta \sim Normal(0,1)$$
$$\sigma \sim Exponential(1)$$

#### 4M4: A sample of students is measured for height each year for 3 years. After the third year, you want to fit a linear regression predicting height using year as a predictor. Write down the mathematical model definition for this regression, using any variable names and priors you choose. Be prepared to defend your choice of priors.

Assume the following model:


$$h_{ij} \sim Normal(\mu_{ij}, \sigma)$$
$$\mu_{ij} = \alpha + \beta(y_i - \bar{y})$$
$$\alpha \sim Normal(68, 10)$$
$$\beta \sim Normal(0,10)$$
$$\sigma \sim Exponential(1)$$
and in R to simulate:

```{r 4M4}
n <- 100
a <- rnorm(n, 68, 10)
b <- rnorm(n, 0, 10)
s <- rexp(n, 1)
y <- 1:3
ybar <- mean(y)
h <- matrix(NA, nrow=n, ncol=3)
for (i in 1:n){
  for (j in 1:3){
    h[i,j] <- rnorm(1, a[i] + b[i]*(y[j] - ybar), s[i])
  }
}
plot(NULL, xlim=c(1,3), ylim=range(h), xlab="Year", ylab="Height (inches)")

for(i in 1:n){
  lines(1:3, h[i,])
}
```

Whoops!  Some students are getting shorter.  

#### 4M5: Now suppose I remind you that every student got taller each year. Does this information lead you to change your choice of priors? How?

Yes, of course.  I messed up with my previous formulation and allowed $\beta$ to be negative.  Whoopsi!  OK, a new method...

$$h_{ij} \sim Normal(\mu_{ij}, \sigma)$$
$$\mu_{ij} = \alpha + \beta(y_i - \bar{y})$$
$$\alpha \sim Normal(68, 10)$$
$$\beta \sim Log-Normal(1,0.5)$$
$$\sigma \sim Exponential(1)$$



```{r 4M5}
n <- 100
a <- rnorm(n, 68, 10)
b <- rlnorm(n, 1, 0.5)
s <- rexp(n, 1)
y <- 1:3
ybar <- mean(y)
h <- matrix(NA, nrow=n, ncol=3)
for (i in 1:n){
  for (j in 1:3){
    h[i,j] <- rnorm(1, a[i] + b[i]*(y[j] - ybar), s[i])
  }
}
plot(NULL, xlim=c(1,3), ylim=range(h), xlab="Year", ylab="Height (inches)")

for(i in 1:n){
  lines(1:3, h[i,])
}


```

#### 4M6: Now suppose I tell you that the variance among heights for students of the same age is never more than 64cm (25 inches). How does this lead you to revise your priors?

$$\sigma \sim Uniform(0,5)$$

because...

$$ \sigma^2 = 25$$
but...

$$\sigma = \sqrt{25} = 5$$

#### 4M7: Refit model m4.3 from the chapter, but omit the mean weight xbar this time. Compare the new model’s posterior to that of the original model. In particular, look at the covariance among the parameters. What is different? Then compare the posterior predictions of both models.

```{r 4M7}
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18,]
m4.3b <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*weight,
    a ~ dnorm(178, 20),
    b ~ dlnorm(0, 1),
    sigma ~ dunif(0, 50)
  ), data = d2
)
precis(m4.3b)
```

Compare with the original model of m4.3.  Notice the differences in intercepts?

```{r}
round(vcov(m4.3b),2)
```
```{r}
round(cov2cor(vcov(m4.3b)),2)
```


#### 4M8: In the chapter, we used 15 knots with the cherry blossom spline. Increase the number of knots and observe what happens to the resulting spline. Then adjust also the width of the prior on the weights—change the standard deviation of the prior and watch what happens. What do you think the combination of knot number and the prior on the weights controls?

As the number of knots increases, the spline will wiggle more but only up to a point.  The width of the prior weights constrains how much the spline can wiggle.  If you make the prior very wide, then the spline can wiggle a lot if it has enough knots.   The combination of number of knots and the width of the prior weights determines how tightly the spline can match the sample.  

### Hard Problems

#### 4H1: The weights listed below were recorded in the !Kung census, but heights were not recorded for these individuals. Provide predicted heights and 89% intervals for each of these individuals. That is, fill in the table below, using model-based predictions.

Use the code as it appears in the chapter.

```{r 4H1}
library(rethinking)
data("Howell1")
d <- Howell1
d2 <- d[d$age >= 18,]
xbar <- mean(d2$weight)

m <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dlnorm(0,1),
    sigma ~ dunif(0,50)
  ), data=d2
)
```

then produce samples from the quadratic approximate posterior.  I used Richard's extract.samples but you could also use mvnorm() just the same.

```{r 4H1cont1}
post <- extract.samples(m)
str(post)
```

now we want to plug the weights in the table into this model and average over the posterior to compute predictions for each individual's height.  

```{r 4H1cont2}
y <- rnorm(1e5, post$a + post$b*(46.95 - xbar), post$sigma)
mean(y)
PI(y, prob=0.89)
```

The code above works by creating a normal distribution sample of 100,000 observations with an assumed weight of 46.95 kg.  We then take the average of that random normal vector and then we get the 89% compatibility interval for height.  

That approach is just for one weight.  We can make this process much easier by using an apply function as follows:

```{r 4H1cont3}
f <- function(weight){
  y <- rnorm(1e5, post$a+ post$b*(46.95 - xbar), post$sigma)
  return(c(mean(y), PI(y,prob=0.89)))
}
weight_list <- c(46.95, 43.72, 64.78, 32.59, 54.63)
result <- sapply(weight_list, f)

## format the output

rtab <- cbind(weight_list, t(result))
colnames(rtab) <- c("weight","height","5%","94%")
rtab
```


#### 4H2: Select out all the rows in the Howell1 data with ages below 18 years of age. If you do it right, you should end up with a new data frame with 192 rows in it.
(a)  Fit a linear regression to these data, using quap. Present and interpret the estimates. For every 10 units of increase in weight, how much taller does the model predict a child gets?

```{r 4H2a}
library(rethinking)
data("Howell1")
d <- Howell1
d3 <- d[d$age < 18,]
str(d3)

xbar <- mean(d3$weight)

m <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(weight - xbar),
    a ~ dnorm(178, 20),
    b ~ dnorm(0,10),
    sigma ~ dunif(0,50)
  ), data=d3
)
precis(m)
```


(b)  Plot the raw data, with height on the vertical axis and weight on the horizontal axis. Superimpose the MAP regression line and 89% interval for the mean. Also superimpose the 89% interval for predicted heights.

```{r 4H2b}
post <- extract.samples(m)
w.seq <- seq(1,45,length.out=50)
mu <- sapply(w.seq, function(z) mean(post$a + post$b*(z-xbar)))
mu.ci <- sapply(w.seq, function(z) PI(post$a + post$b*(z-xbar),prob=0.89))
pred.ci <- sapply(w.seq, function(z) PI(rnorm(1e4, post$a + post$b*(z-xbar),post$sigma),prob=0.89))

plot(height ~ weight, data=d3, col=col.alpha("slateblue",0.5), cex=.5)
lines(w.seq, mu)
lines(w.seq, mu.ci[1,], lty=2)
lines(w.seq, mu.ci[2,], lty=2)
lines(w.seq, pred.ci[1,], lty=2)
lines(w.seq, pred.ci[2,], lty=2)
```

(c)  What aspects of the model fit concern you? Describe the kinds of assumptions you would change, if any, to improve the model. You don’t have to write any new code. Just explain what the model appears to be doing a bad job of, and what you hypothesize would be a better model.

Weight and height are not linear.  Look carefully.  

#### 4H3: Suppose a colleague of yours, who works on allometry, glances at the practice problems just above. Your colleague exclaims, “That’s silly. Everyone knows that it’s only the logarithm of body weight that scales with height!” Let’s take your colleague’s advice and see what happens.

(a)  Model the relationship between height (cm) and the natural logarithm of weight (log-kg). Use the entire Howell1 data frame, all 544 rows, adults and non-adults. Can you interpret the resulting estimates?

```{r 4H3a}
library(rethinking)
data("Howell1")
d <- Howell1
logxbar <- mean(log(d$weight))

mlw <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b*(log(weight) - logxbar),
    a ~ dnorm(178, 20),
    b ~ dnorm(0,10),
    sigma ~ dunif(0,50)
  ), data=d
)
precis(mlw)
```



(b)  Begin with this plot: plot( height ~ weight , data=Howell1 ). Then use samples from the quadratic approximate posterior of the model in (a) to superimpose on the plot: (1) the predicted mean height as a function of weight, (2) the 97% interval for the mean, and (3) the 97% interval for predicted heights.

```{r 4H3b}

post <- extract.samples(mlw)
w.seq <- seq(4,63, length.out=50)
mu <- sapply(w.seq, function(z) mean(post$a+post$b*(log(z)-logxbar)))
mu.ci <- sapply(w.seq, function(z) PI(post$a+post$b*(log(z)-logxbar)))
h.ci <- sapply(w.seq, function(z) PI(rnorm(1e4,post$a+post$b*(log(z)-logxbar),post$sigma)))

plot(height ~ weight, data=d, col=col.alpha("slateblue",0.5), cex=.5)
lines(w.seq, mu)
lines(w.seq, mu.ci[1,], lty=2)
lines(w.seq, mu.ci[2,], lty=2)
lines(w.seq, pred.ci[1,], lty=2)
lines(w.seq, pred.ci[2,], lty=2)

```



#### 4H4: Plot the prior predictive distribution for the parabolic polynomial regression model in the chapter. You can modify the code that plots the linear regression prior predictive distribution. Can you modify the prior distributions of α, β1, and β2 so that the prior predictions stay within the biologically reasonable outcome space? That is to say: Do not try to fit the data by hand. But do try to keep the curves consistent with what you know about height and weight, before seeing these exact data.

```{r 4H4}
library(rethinking)
data("Howell1")
d <- Howell1

d$weight_s <- scale(d$weight)
d$weight_s2 <- d$weight_s^2

m4.5 <- quap(
  alist(
    height ~ dnorm(mu, sigma),
    mu <- a + b1*weight_s + b2*weight_s2,
    a ~ dnorm(178, 20),
    b1 ~ dlnorm(0,1),
    b2 ~ dnorm(0,1),
    sigma ~ dunif(0,50)
  ), data=d
)
set.seed(45)
prior <- extract.prior(m4.5)
precis(prior)
```


#### 4H5: Return to data(cherry_blossoms) and model the association between blossom date (doy) and March temperature (temp). Note that there are many missing values in both variables. You may consider a linear model, a polynomial, or a spline on temperature. How well does temperature trend predict the blossom trend?

```{r 4H5}
library(rethinking)
data(cherry_blossoms)
colSums( is.na(cherry_blossoms) )
d <- cherry_blossoms
d2 <- d[ complete.cases( d$doy , d$temp ) , c("doy","temp") ]

num_knots <- 30
knot_list <- quantile( d2$temp , probs=seq(0,1,length.out=num_knots) )
library(splines)
B <- bs(d2$temp,
knots=knot_list[-c(1,num_knots)] ,
degree=3 , intercept=TRUE )

m4H5 <- quap(
  alist(
    D ~ dnorm( mu , sigma ) ,
    mu <- a + B %*% w ,
    a ~ dnorm(100,10),
    w ~ dnorm(0,10),
    sigma ~ dexp(1)
    ), data=list( D=d2$doy , B=B ) ,
  start=list( w=rep( 0 , ncol(B) ) ) )

mu <- link( m4H5 )
mu_mean <- apply( mu , 2 , mean )
mu_PI <- apply( mu , 2 , PI, 0.97 )
plot( d2$temp , d2$doy , col=col.alpha(rangi2,0.3) , pch=16 ,
xlab="temp" , ylab="doy" )
o <- order( d2$temp )
lines( d2$temp[o] , mu_mean[o] , lwd=3 )
shade( mu_PI[,o] , d2$temp[o] , col=grau(0.3) )
```


#### 4H6:  Simulate the prior predictive distribution for the cherry blossom spline in the chapter. Adjust the prior on the weights and observe what happens. What do you think the prior on the weights is doing?

```{r 4H6}
library(rethinking)
data(cherry_blossoms)
d <- cherry_blossoms
d2 <- d[ complete.cases(d$doy) , ] # complete cases on doy
num_knots <- 15
knot_list <- quantile( d2$year , probs=seq(0,1,length.out=num_knots) )
library(splines)
B <- bs(d2$year,
knots=knot_list[-c(1,num_knots)] ,
degree=3 , intercept=TRUE )
m4.7 <- quap(
alist(
D ~ dnorm( mu , sigma ) ,
mu <- a + B %*% w ,
a ~ dnorm(100,10),
w ~ dnorm(0,10),
sigma ~ dexp(1)
), data=list( D=d2$doy , B=B ) ,
start=list( w=rep( 0 , ncol(B) ) ) )

p <- extract.prior( m4.7 )
mu <- link( m4.7 , post=p )
mu_mean <- apply( mu , 2 , mean )
mu_PI <- apply( mu , 2 , PI, 0.97 )
plot( d2$year , d2$doy , col=col.alpha(rangi2,0.3) , pch=16 ,
xlab="year" , ylab="doy" , ylim=c(60,140) )
lines( d2$year , mu_mean , lwd=3 )
shade( mu_PI , d2$year , col=grau(0.3) )
p <- extract.prior( m4.7 )
mu <- link( m4.7 , post=p )
plot( d2$year , d2$doy , col=col.alpha(rangi2,0.3) , pch=16 ,
xlab="year" , ylab="day in year" , ylim=c(60,140) )
for ( i in 1:20 ) lines( d2$year , mu[i,] , lwd=1 )
m4.7b <- quap(
alist(
D ~ dnorm( mu , sigma ) ,
mu <- a + B %*% w ,
a ~ dnorm(100,10),
w ~ dnorm(0,1),
sigma ~ dexp(1)
), data=list( D=d2$doy , B=B ) ,
start=list( w=rep( 0 , ncol(B) ) ) )
p <- extract.prior( m4.7b )
mu <- link( m4.7b , post=p )
plot( d2$year , d2$doy , col=col.alpha(rangi2,0.3) , pch=16 ,
xlab="year" , ylab="day in year" , ylim=c(60,140) )
for ( i in 1:20 ) lines( d2$year , mu[i,] , lwd=1 )
```


#### 4H7: TYPO - no 7th Hard problem

#### 4H8: The cherry blossom spline in the chapter used an intercept α, but technically it doesn’t require one. The first basis functions could substitute for the intercept. Try refitting the cherry blossom spline without the intercept. What else about the model do you need to change to make this work?

```{r 4H8}
library(rethinking)
data(cherry_blossoms)
d <- cherry_blossoms
d2 <- d[ complete.cases(d$doy) , ] # complete cases on doy
num_knots <- 15
knot_list <- quantile( d2$year , probs=seq(0,1,length.out=num_knots) )
library(splines)
B <- bs(d2$year,
knots=knot_list[-c(1,num_knots)] ,
degree=3 , intercept=TRUE )
m4H7 <- quap(
alist(
D ~ dnorm( mu , sigma ) ,
mu <- 0 + B %*% w ,
w ~ dnorm(100,10),
sigma ~ dexp(1)
), data=list( D=d2$doy , B=B ) ,
start=list( w=rep( 0 , ncol(B) ) ) )
```


# Chapter 5

$$A \perp \!\!\! \perp B $$
Some key terms to consider:
* spurious relationships
* confounded relationships
* multicollinearity

Much of the chapter deals with testing models using a classical nested model comparison approach but applied to Bayesian models.  Be sure to get the gist of the chapter before trying to dig deep into the details or homework problems.  

## Useful functions to know in the rethinking package

* link()
* sim()
* PI()
* HPDI()
* extract.prior()
* precis()

## Useful packages that are either used in the chapter or that I mentioned in class

* [dagitty](https://cran.r-project.org/web/packages/dagitty/index.html)
* [Rgraphviz](https://www.bioconductor.org/packages/release/bioc/html/Rgraphviz.html)

I prefer the latter but I am also more familiar with standard SEM notation than I am with the rather sparse DAG notation.

Let's discuss these options further so you know what they do and why they might be useful even with STAN code and shiny-stan.

## Assigned Problems

### Easy Problems

#### 5E1:  Which of the linear models below are multiple linear regressions?

Models 2 and 4.  Should be pretty obvious now.

#### 5E2:  Write down a multiple regression to evaluate the claim: Animal diversity is linearly related to latitude, but only after controlling for plant diversity. You just need to write down the model definition.

$$D_i \sim Normal(\mu_i, \sigma) $$
$$\mu_i = \alpha + \beta_L L_i + \beta_P P_i $$

#### 5E3:  Write down a multiple regression to evaluate the claim: Neither amount of funding nor size of laboratory is by itself a good predictor of time to PhD degree; but together these variables are both positively associated with time to degree. Write down the model definition and indicate which side of zero each slope parameter should be on.

$$P_i \sim Normal(\mu_i, \sigma)$$
$$\mu_i = \alpha + \beta_F F_i + \beta_S S_i$$

with both $\beta_F$ and $\beta_S$ non-zero but negligible.  Where the question can be a bit difficult is interpreting the independence of funding and size of lab are negligible but together they form a positive prediction.  If they are independent but together produce a strong predictive model then we can see both $\beta$'s being small but not necessarily significant while the total predictive validity of the model is sufficient to justify its use (i.e., small residuals with no outliers).  Even if they are independent, we could postulate an interaction (something we have yet to discuss in the class).  An interaction would indicate that neither alone are important but together they form a composite effect that reduces the residuals in our prediction model.  Think of this moderated effect as an alternative variable produced by multiplying the two predictors and, as such, requiring the specification of another $\beta_{FxS}$ that would account for the interaction.

#### 5E4:  Suppose you have a single categorical predictor with 4 levels (unique values), labeled A, B, C and D. Let Ai be an indicator variable that is 1 where case i is in category A. Also suppose Bi, Ci, and Di for the other categories. Now which of the following linear models are inferentially equivalent ways to include the categorical variable in a regression? Models are inferentially equivalent when it’s possible to compute one posterior distribution from the posterior distribution of another model.

Richard says that models 1, 3, 4, and 5 are inferentially equivalent with weak priors.  I'm not sure I agree on that point but we do agree that model 2 has a redundant parameter ($\alpha$ and $\beta_A$)  Here is my thinking on the other four.  Model 1 omits that C influence and, as such, has the C contribution in the intercept.  Model 3 omits A and, as such, has the A contribution in the intercept.  Similarly, model 5 just makes explicit model 3 so I have no problem with that one.  Where I do have a problem is in the explicit notation for model 4.  That model either has an explicitly stated intercept for A and, as such, is the same model as model 3 OR it is a typo.  I'll ask Richard directly and get back to you.

### Medium Problems

#### 5M1: Invent your own example of a spurious correlation. An outcome variable should be correlated with both predictor variables. But when both predictors are entered in the same model, the correlation between the outcome and one of the predictors should mostly vanish (or at least be greatly reduced).

[Read This Article For A Great Example](https://fivethirtyeight.com/features/science-isnt-broken/#part1)


#### 5M2: Invent your own example of a masked relationship. An outcome variable should be correlated with both predictor variables, but in opposite directions. And the two predictor variables should be correlated with one another.

Consider divorce differentially affecting males compared to females.  Why?  I don't need to invent anything here.  Also, if you have not heard of the purported effect, the [Dunning-Kruger effect](https://github.com/pem725/Dunning-Kruger.git) is one that masks a relationship - the true relationship.  I have more.  Let's discuss others.

#### 5M3: It is sometimes observed that the best predictor of fire risk is the presence of firefighters—States and localities with many firefighters also have more fires. Presumably firefighters do not cause fires. Nevertheless, this is not a spurious correlation. Instead fires cause firefighters. Consider the same reversal of causal inference in the context of the divorce and marriage data. How might a high divorce rate cause a higher marriage rate? Can you think of a way to evaluate this relationship, using multiple regression?

Yes.  Time.  Everyone omits time in social science.  If you modeled the relationship including time, you could see that firefighters come after fires.  The omission of a relevant variable (here, time) is deadly in science and scientific inference.

#### 5M4:  In the divorce data, States with high numbers of members of the Church of Jesus Christ of Latter-day Saints (LDS) have much lower divorce rates than the regression models expected. Find a list of LDS population by State and use those numbers as a predictor variable, predicting divorce rate using marriage rate, median age at marriage, and percent LDS population (possibly standardized). You may want to consider transformations of the raw percent LDS variable.

```{r 5M4}
library(rethinking)
library(rvest)
library(tidyverse)

data(WaffleDivorce)
d <- WaffleDivorce

## scrape from wikipedia
LDSpage <- read_html("https://en.wikipedia.org/wiki/The_Church_of_Jesus_Christ_of_Latter-day_Saints_membership_statistics_(United_States)")

LDStab <- LDSpage %>% 
  html_nodes(css = "table") %>%
  nth(4) %>%
  html_table(fill=TRUE)

LDStab <- LDStab[,c(1,4)]
LDStab$LDS <- as.numeric(gsub("[\\%,]", "",LDStab$LDS))
d <- merge(d,LDStab,by.x="Location",by.y="State",all.x=T)

d$L <- standardize( d$LDS )
d$A <- standardize( d$MedianAgeMarriage )
d$M <- standardize( d$Marriage )
d$D <- standardize( d$Divorce )

m_5M4 <- quap(
  alist(
    D ~ dnorm(mu,sigma),
    mu <- a + bM*M + bA*A + bL*L,
    a ~ dnorm(0,0.2),
    c(bA,bM,bL) ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )
precis( m_5M4 )

```


#### 5M5: One way to reason through multiple causation hypotheses is to imagine detailed mechanisms through which predictor variables may influence outcomes. For example, it is sometimes argued that the price of gasoline (predictor variable) is positively associated with lower obesity rates (outcome variable). However, there are at least two important mechanisms by which the price of gas could reduce obesity. First, it could lead to less driving and therefore more exercise. Second, it could lead to less driving, which leads to less eating out, which leads to less consumption of huge restaurant meals. Can you outline one or more multiple regressions that address these two mechanisms? Assume you can have any predictor data you need.

This question is an exercise in your knowledge and creativity.  While I think it is always a good idea to get students to think about these topics, most lack the broad knowledge to capture what Richard is asking here.  You need to think about alternative causal models where the DAG is almost certainly wrong.  These may be in short supply for most of you so I will provide a few suggestions on Thursday when we meet again.


### Hard Problems

#### 5H1: In the divorce example, suppose the DAG is: M → A → D. What are the implied conditional independencies of the graph? Are the data consistent with it?

```{r 5H1}
library(dagitty)
dag_5H1 <- dagitty("dag{M->A->D}")
impliedConditionalIndependencies(dag_5H1)
```


#### 5H2: Assuming that the DAG for the divorce example is indeed M → A → D, fit a new model and use it to estimate the counterfactual effect of halving a State’s marriage rate M. Use the counterfactual example from the chapter (starting on page 140) as a template.

```{r 5H2}
library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce
d$D <- standardize( d$Divorce )
d$M <- standardize( d$Marriage )
d$A <- standardize( d$MedianAgeMarriage )

m5H2 <- quap(
  alist(
    # A -> D
    D ~ dnorm( muD , sigmaD ),
    muD <- aD + bAD*A,
    # M -> A
    A ~ dnorm( muA , sigmaA ),
    muA <- aA + bMA*M,
    # priors
    c(aD,aA) ~ dnorm(0,0.2),
    c(bAD,bMA) ~ dnorm(0,0.5),
    c(sigmaD,sigmaA) ~ dexp(1)
    ) , data=d )
precis(m5H2)
M_seq <- seq( from=-3 , to=3 , length.out=30 )
sim_dat <- data.frame( M=M_seq )
s <- sim( m5H2 , data=sim_dat , vars=c("A","D") )

plot( sim_dat$M , colMeans(s$A) , ylim=c(-2,2) , type="l" ,xlab="manipulated M" ,
      ylab="counterfactual A" )
shade( apply(s$A,2,PI) , sim_dat$M )
mtext( "Counterfactual M -> A" )

plot( sim_dat$M , colMeans(s$D) , ylim=c(-2,2) , type="l" ,
      xlab="manipulated M" , ylab="counterfactual D" )
shade( apply(s$D,2,PI) , sim_dat$M )
mtext( "Counterfactual M -> A -> D" )
(10 - mean(d$Marriage))/sd(d$Marriage)
M_seq <- c( 0 , -2.67 )
sim_dat <- data.frame( M=M_seq )
s <- sim( m5H2 , data=sim_dat , vars=c("A","D") )
diff <- s$D[,2] - s$D[,1]
mean( diff )
```


#### 5H3: Return to the milk energy model, m5.7. Suppose that the true causal relationship among the variables is: (see diagram in book).  Now compute the counterfactual effect on K of doubling M. You will need to account for both the direct and indirect paths of causation. Use the counterfactual example from the chapter (starting on page 140) as a template.

```{r 5H3}
library(rethinking)
data(milk)
d <- milk
d$K <- standardize( d$kcal.per.g )
d$N <- standardize( d$neocortex.perc )
d$M <- standardize( log(d$mass) )
d2 <- d[ complete.cases( d$K , d$N , d$M ) , ]
m5H3 <- quap(
  alist(
    # M -> K <- N
    K ~ dnorm( muK , sigmaK ),
    muK <- aK + bMK*M + bNK*N,
    # M -> N
    N ~ dnorm( muN , sigmaN ),
    muN <- aN + bMN*M,
    # priors
    c(aK,aN) ~ dnorm( 0 , 0.2 ),
    c(bMK,bNK,bMN) ~ dnorm( 0 , 0.5 ),
    c(sigmaK,sigmaN) ~ dexp( 1 )
    ) , data=d2 )
precis( m5H3 )
(log(c(15,30)) - mean(log(d$mass)))/sd(log(d$mass))
M_seq <- c( 0.75 , 1.15 )
sim_dat <- data.frame( M=M_seq )
s <- sim( m5H3 , data=sim_dat , vars=c("N","K") )
diff <- s$K[,2] - s$K[,1]
quantile( diff , probs=c( 0.05 , 0.5 , 0.94 ) )
M_seq <- seq( from=-3 , to=3 , length.out=30 )
sim_dat <- data.frame( M=M_seq )
s <- sim( m5H3 , data=sim_dat , vars=c("N","K") )
0.61*0.68 - 0.7
( 0.61*0.68 - 0.7 )*0.4

```


#### 5H4: Here is an open practice problem to engage your imagination. In the divorce date, States in the southern United States have many of the highest divorce rates. Add the South indicator variable to the analysis. First, draw one or more DAGs that represent your ideas for how Southern American culture might influence any of the other three variables (D, M or A). Then list the testable implications of your DAGs, if there are any, and fit one or more models to evaluate the implications. What do you think the influence of “Southerness” is?

```{r 5H4}
library(dagitty)
dag_5H4 <- dagitty("dag{ M <- A -> D; A <- S -> D; M -> D }")
coordinates(dag_5H4) <- list(
  x=c(A=0,M=1,D=2,S=1),
  y=c(A=1,M=0,D=1,S=2) )
drawdag( dag_5H4 )
impliedConditionalIndependencies( dag_5H4 )

library(rethinking)
data(WaffleDivorce)
d <- WaffleDivorce
d$D <- standardize( d$Divorce )
d$M <- standardize( d$Marriage )
d$A <- standardize( d$MedianAgeMarriage )
d$S <- d$South
m_5H4 <- quap(
  alist(
    M ~ dnorm( mu , sigma ),
    mu <- a + bS*S + bA*A,
    a ~ dnorm( 0 , 0.2 ),
    c(bS,bA) ~ dnorm( 0 , 0.5 ),
    sigma ~ dexp( 1 )
    ) , data=d )
precis( m_5H4 )

```

# Chapter 6

The chapter focuses on what happens when causal relationships are ignored (or not tested) but inferred from regression models.  Richard cites several biases or paradoxes that we may fall prey to without a suitable understanding of causality.

* [Berkson's Paradox:](https://en.wikipedia.org/wiki/Berkson%27s_paradox) A selection bias problem that seems to have caught Richard's eye.  The paradox has some relevance to social science but not much.  
* Collider Bias: see above
* Multicollinearity (see above), non-identifiability, and included variable bias
* [Simpson's paradox:](https://en.wikipedia.org/wiki/Simpson%27s_paradox)
* Conditioning on the consequence rather than the antecedent (another Collider bias)

DAGs may be helpful in getting us to root out confounds.  Richard provides four elemental ones - the Fork, Pipe, Collider, and Descendent.

## The Fork



```{r ForkDAG, fig.cap="Fork DAG"}
library(dagitty)
fork_dag <- dagitty('dag {
  Z [pos="0,0"]
  X [pos="-1,-1"]
  Y [pos="1,-1"]
  
  Z -> {X Y}
}')
plot(fork_dag)

```

The *FORK* is the classic confounder where a variable Z is a common cause of X and Y and, as such, generates a correlation between them. Conditioning our effects on Z, then X tells us nothing about Y. X and Y are independent, conditional on Z.

```{r}
impliedConditionalIndependencies(fork_dag)
```

## The Pipe

```{r PipeDAG, fig.cap="Pipe DAG"}
pipe_dag <- dagitty('dag {
  X [pos="-1,-1"]
  Z [pos="0,0"]
  Y [pos="1,1"]
  
  X -> Z
  Z -> Y
}')
plot(pipe_dag)

```

The treatment X influences Z which influences Y - a pure mediation effect. If we condition on Z, we block the path from X to Y. Conditioning on the middle variable in both the **FORK** and the **PIPE** blocks the path between X and Y.

```{r}
impliedConditionalIndependencies(pipe_dag)
```
## The Collider

```{r ColliderDAG, fig.cap="Collider DAG"}
collider_dag <- dagitty('dag {
  X [pos="-1,1"]
  Z [pos="0,0"]
  Y [pos="1,1"]
  
  X -> Z
  Y -> Z
}')
plot(collider_dag)

```

There is no association between X and Y in a **COLLIDER** unless you condition on Z. Conditioning on Z (aka, the collider variable), opens the path between X and Y.  Simply put, once the path is open, information flows between X and Y, however, neither X nor Y has any causal influence on the other.

```{r}
impliedConditionalIndependencies(collider_dag)
```


## The Descendent

```{r DescendentDAG, fig.cap="Descendent DAG"}
descendent_dag <- dagitty('dag {
  X [pos="-1,1"]
  Z [pos="0,0"]
  Y [pos="1,1"]
  D [pos="0,.7"]
  
  X -> Z
  Y -> Z
  Z -> D
}')
plot(descendent_dag)

```

A descendent (D) is a variable influenced by another variable (Z). Conditioning on a descendent partly conditions on its parent because D has *some* information about Z but the information is incomplete (often in the form of a proxy for Z). Further, conditioning on D partially opens the path from X to Y, because Z is a collider. Note that the effects of conditioning may change depending upon the nature of its parents with respect to X and Y.  

```{r}
impliedConditionalIndependencies(descendent_dag)
```
Be sure to run the code that Richard provided in the chapter and make sure you can reproduce the effects.

## Assigned Problems

### Easy Problems

#### 6E1.  List three mechanisms by which multiple regression can produce false inferences about causal effects.

1. Confounding
2. Multicollineary
3. Conditioning on Post-Tx variable
4. Collider bias

#### 6E2.  For one of the mechanisms in the previous problem, provide an example of your choice, perhaps from your own research.

Math ability and shoe size

```{r 6E2}
library(dagitty)
ans6E2 <- dagitty('dag {
  Age -> Education
  Age -> ShoeSize
  Education -> MathAbility
}')
plot(ans6E2)
impliedConditionalIndependencies(ans6E2)
```



#### 6E3.  List the four elemental confounds. Can you explain the conditional dependencies of each?

1. Fork (see above for all)
2. Pipe
3. Collider
4. Descendent


#### 6E4.  How is a biased sample like conditioning on a collider? Think of the example at the open of the chapter.

A biased sample means that there is necessarily a restriction of range between X and Y.  That restriction can come in the form of many effects but a collider is one of the many.  The bias may result in an unfortunate conditioning that is like a collider but it may come in the form of the other confounds (fork, pipe, or descendent).

### Medium Problems

#### 6M1.  Modify the DAG on page 186 to include the variable V, an unobserved cause of C and Y: C ← V → Y. Reanalyze the DAG. How many paths connect X to Y? Which must be closed? Which variables should you condition on now?

```{r 6M1} 
library(dagitty)
dag_6M1 <- dagitty("dag{
  U [unobserved]
  V [unobserved]
  X -> Y
  X <- U -> B <- C -> Y
  U <- A -> C
  C <- V -> Y }")

coordinates(dag_6M1) <- list(
  x=c(X=0,Y=2,U=0,A=1,B=1,C=2,V=2.5),
  y=c(X=2,Y=2,U=1,A=0.5,B=1.5,C=1,V=1.5) )
plot(dag_6M1)
```

Given the 5 paths connecting X and Y:
1. X -> Y (the direct causal path)
2. X <- U -> B <- C -> Y
3. X <- U -> B <- C <- V -> Y
4. X <- U <- A -> C -> Y
5. X <- U <- A -> C <- V -> Y

Path 1 should remain open and close the others because they are non-causal (i.e., confounds).  The B paths are closed since B is a collider so we don't condition on B.  Paths through A vna V are also closed because C is a collider on them.  We can condition on A to close all the non-causal paths and demonstrate this by the following code:

```{r 6M1cont1}
adjustmentSets(dag_6M1, exposure = "X", outcome="Y")

```
The chapter showed that we could also condition on C when V was not included but with V, we cannot condition on C.

```{r 6M1cont2} 
library(dagitty)
dag186 <- dagitty("dag{
  U [unobserved]
  X -> Y
  X <- U -> B <- C -> Y
  U <- A -> C}")

coordinates(dag186) <- list(
  x=c(X=0,Y=2,U=0,A=1,B=1,C=2),
  y=c(X=2,Y=2,U=1,A=0.5,B=1.5,C=1) )
plot(dag186)
```

```{r dag186}
adjustmentSets(dag186, exposure = "X", outcome="Y")
```
See?

#### 6M2.  Sometimes, in order to avoid multicollinearity, people inspect pairwise correlations among predictors before including them in a model. This is a bad procedure, because what matters is the conditional association, not the association before the variables are included in the model. To highlight this, consider the DAG X → Z → Y. Simulate data from this DAG so that the correlation between X and Z is very large. Then include both in a model prediction Y. Do you observe any multicollinearity? Why or why not? What is different from the legs example in the chapter?

```{r 6M2.1}
N <- 1000
X <- rnorm(N)
Z <- rnorm(N,X,0.1)
Y <- rnorm(N,Z)
cor(X,Z)
```

```{r 6M2.2}
m_6M2 <- quap(
  alist(
    Y ~ dnorm( mu , sigma ),
    mu <- a + bX*X + bZ*Z,
    c(a,bX,bZ) ~ dnorm(0,1),
    sigma ~ dexp(1)
    ) , data=list(X=X,Y=Y,Z=Z) )
precis( m_6M2 )

```

Conditioning on a post-treatment variable because it eliminates X since Z mediates the entire effect between X and Y.  


#### 6M3.  Learning to analyze DAGs requires practice. For each of the four DAGs below, state which variables, if any, you must adjust for (condition on) to estimate the total causal influence of X on Y.

In each case, the procedure is the same: List all paths between X and Y and figure out which non-causal paths are open and which variables you can use to close them. We’ll consider each in turn.

Top left. There are three paths between X and Y: (1) X → Y, (2) X ← Z → Y, (3) X ← Z ← A → Y. Both (2) and (3) are open, non-causal paths. Conditioning on Z is sufficient to close both.

Top right. Again three paths: (1) X → Y, (2) X → Z → Y, (3) X → Z ← A → Y. Paths (1) and (2) are both causal. We want both open. Path (3) is non-causal, but it is also closed already because Z is a collider on that path.

Bottom left. Paths: (1) X → Y, (2) X → Z ← Y, (3) X ← A → Z ← Y. Only path (1) is causal. The other paths are both closed, because both contain a collider at Z. Conditioning on Z would be a disaster.

Bottom right. Paths: (1) X → Y, (2) X → Z → Y, (3) X ← A → Z → Y. Paths (1) and (2) are causal. Path (3) is an open backdoor path. To close it, we must condition on A or Z, but if condition on Z, that would close a causal path.

### Hard Problems

#### 6H1.  Use the Waffle House data, data(WaffleDivorce), to find the total causal influence of number of Waffle Houses on divorce rate. Justify your model or models with a causal graph.

```{r 6H1}
library(dagitty)
dag_6H1 <- dagitty("dag{A -> D; A -> M -> D; A <- S -> D; S -> W -> D}")
coordinates(dag_6H1) <- list(
  x=c(A=0,M=1,D=2,S=1,W=2),
  y=c(A=0.75,M=0.5,D=1,S=0,W=0) )
plot(dag_6H1)
```

```{r 6H1cont}
adjustmentSets(dag_6H1, exposure = "A", outcome="D")
```

#### 6H2.  Build a series of models to test the implied conditional independencies of the causal graph you used in the previous problem. If any of the tests fail, how do you think the graph needs to be amended? Does the graph need more or fewer arrows? Feel free to nominate variables that aren’t in the data.

```{r 6H2}
impliedConditionalIndependencies( dag_6H1 )
```

All three problems below are based on the same data. The data in data(foxes) are 116 foxes from 30 different urban groups in England. These foxes are like street gangs. Group size varies from 2 to 8 individuals. Each group maintains its own urban territory. Some territories are larger than others. The area variable encodes this information. Some territories also have more avgfood than others. We want to model the weight of each fox. For the problems below, assume the following DAG:

#### 6H3.  Use a model to infer the total causal influence of area on weight. Would increasing the area available to each fox make it heavier (healthier)? You might want to standardize the variables. Regardless, use prior predictive simulation to show that your model’s prior predictions stay within the possible outcome range.

```{r 6H3}
library(rethinking)
data(foxes)
d <- foxes
d$W <- standardize(d$weight)
d$A <- standardize(d$area)
m1 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bA*A,
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )
precis(m1)
```

Territory size seems to have no total causal influence on weight, at least not in this sample.

#### 6H4.  Now infer the causal impact of adding food to a territory. Would this make foxes heavier? Which covariates do you need to adjust for to estimate the total causal influence of food?

To infer the causal influence of avgfood on weight, we need to close any back-door paths. There are no back-door paths in the DAG. So again, just use a model with a single predictor. If you include groupsize, to block the indirect path, then you won’t get the total causal influence of food. You’ll just get the direct influence. But I asked for the effect of adding food, and that would mean through all forward paths.

```{r 6H4}
d$F <- standardize(d$avgfood)
m2 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bF*F,
    a ~ dnorm(0,0.2),
    bF ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )
precis(m2)
```

Again nothing. Adding food does not change weight. This shouldn’t surprise you, if the DAG is correct, because area is upstream of avgfood.

#### 6H5.  Now infer the causal impact of group size. Which covariates do you need to adjust for? Looking at the posterior distribution of the resulting model, what do you think explains these data? That is, can you explain the estimates for all three problems? How do they go together?

The variable groupsize does have a back-door path, passing through avgfood. So to infer the causal influence of groupsize, we need to close that path. This implies a model with both groupsize and avgfood as predictors.

```{r 6H5}
d$G <- standardize(d$groupsize)
m3 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bF*F + bG*G,
    a ~ dnorm(0,0.2),
    c(bF,bG) ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )
precis(m3)
```

It looks like group size is negatively associated with weight, controlling for food. Similarly, food is positively associated with weight, controlling for group size. So the causal influence of group size is to reduce weight—less food for each fox. And the direct causal influence of food is positive, of course. But the total causal influence of food is still nothing, since it causes larger groups. This is a masking effect, like in the milk energy example. But the causal explanation here is that more foxes move into a territory until the food available to each is no better than the food in a neighboring territory. Every territory ends up equally good/bad on average. This is known in behavioral ecology as an ideal free distribution. 

#### 6H6.  Consider your own research question. Draw a DAG to represent it. What are the testable implications of your DAG? Are there any variables you could condition on to close all backdoor paths? Are there unobserved variables that you have omitted? Would a reasonable colleague imagine additional threats to causal inference that you have ignored?

Good exercise for each of you.  Show me what you got.

#### 6H7.  For the DAG you made in the previous problem, can you write a data generating simulation for it? Can you design one or more statistical models to produce causal estimates? If so, try to calculate interesting counterfactuals. If not, use the simulation to estimate the size of the bias you might expect. Under what conditions would you, for example, infer the opposite of a true causal effect?

Use the tools I showed you from the above.  Now is a good time to try these tools out in a real applied problem that you find interesting/engaging.


# Chapter 7

## Points to remember:

* Two incorrect models may produce similar results
* Parsimony - fewer assumptions and fewer parameters - leads the way!
* Beware of the overfit and underfit models - both mislead
* Two approaches to defend against these models
  - Use a **regularizing prior**
  - Use a scoring device like an **information criterion** or **cross-validation**
* A word of caution - information criteria are tough to understand but easy to implement
* NOTE:  Confounded models *may* outperform correct causal models
* Please read and re-read the Rethinking: OLS and Bayesian anti-essentialism.  You will learn about the ties between frequentists and subjectivists at a deeper level.
* **OVERFITTING:**  As the number of parameters increases, we learn less and less about the data.  Richard refers to this as **data compression** but it is actually a form of data reduction (cf EFA, CFA, PCA).  A model completely expressing the observed data is referred to as the MDL or Mimimum Description Length.  Avoid this outcome!  Too many parameters means you trust your data more than the model leading to poor predictive validity in the future.
* **UNDERFITTING:** Too few parameters means you trust the model more than your data.  Beware of the restricted model where you may be inclined to let parsimony reign over the kingdom of scientific prediction.
* Accuracy of model predictions:
  - Cost-benefit. Assign weights to decisions and compute costs and benefits accordingly.
  - Accuracy in context.  Compute scoring rules on accuracy such as the *Log Scoring Rule*.
* Calibration does not equate to accuracy.  NOTE:  The D-K effect is one that clearly shows this problem in detail.

## Data are NOT information

A classical problem that people often make is that data is information; it is not.  Data are information *only* when the data lead to a decrease in uncertainty.  I want you all to embrace this clear definition because, by doing so, you will be a more effective data analyst.  Further, information and the pursuit of reducing uncertainty is the fundamental aspect of Bayesian data analysis.  

## Uncertainty Desiderata

1. Continuous as opposed to discrete
2. Directly proportional to the number of possible events
3. Additive (i.e., conjoint additivity conditions must be met)

## Information

$$H(p) = -E \; log(p_i) = - \sum_{i=1}^{n} p_i log(p_i) $$

**Entropy** $H(p)$ is *"the uncertainty contained in a probability distribution is the average log-probability of an event."*

Be sure you understand this concept because entropy is the key bit to evaluating Bayesian model results.  Computation of entropy is quite simply.  If you have two discrete events (total possible events are 2) and the probabilities are known (in retrospect or a prior hypothesized), we can compute entropy $H(p)$.

```{r EntropyDemoCalc}
p <- c(.2,.8)
round(-sum(p*log(p)),2)
```

Consider how two discrete and, as a result, ipsative events (i.e., mutually exclusive) can be plotted with respect to the probability of one.

```{r EntropyFig, warning=FALSE, message=FALSE}
library(ggplot2)
tmp <- data.frame(p1=seq(0,1,length=100))
tmp$p2 <- 1-tmp$p1

binH <- function(x){
  -sum(x*log(x))
}

tmp$H <- apply(tmp,1,binH)
ggplot(tmp,aes(p1,H)) + geom_smooth() + ylab("Entropy (H)") + xlab("Probability of P1")

```

Entropy is maximal when the probability is even at 0.5.  What does that mean?  Think about it.  Maximum uncertainty is best represented as a coin toss.  We spent most of the past 6 weeks talking about priors and uncertainty about priors.  In each of these operations, we specified a uniform prior (i.e., every observation had an equal probability of occurring).  A discrete, binary event has two outcomes and if those outcomes are equal, we are at our maximum uncertainty when $p=0.5$.  

Bayesian updating maximizes entropy (MAXENT).  Think about it.  

In contrast to entropy where our focus is largely on uncertainty, we have divergence $D_{KL}$ that provides us with the average difference in log probabilities between the target (p) and model (q).  Divergence, in short, is a way for us to quantify our accuracy.  We formalize the definition of divergence using the following mathematical notation:

$$D_{KL}(p,q) = \sum_i p_i(log(p_i) - log(q_i)) = \sum_i p_i log(\frac{p_i}{q_i}) $$
**Divergence** is *"the additional uncertainty induced by using probabilities from one distribution to describe another distribution."*  We use divergence to estimate the distance from our model (q) to reality (p) or, more generally between any two models.  

## Deviance and Simulation

```{r SimTrainTest, eval=FALSE}
library(rethinking)
N <- 20
kseq <- 1:5
dev <- sapply( kseq , function(k) {
  print(k);
  r <- replicate( 1e4 , sim_train_test( N=N, k=k ),mc.cores=20 );
  c( mean(r[1,]) , mean(r[2,]) , sd(r[1,]) , sd(r[2,]) )
} )

plot( 1:5 , dev[1,] , ylim=c( min(dev[1:2,])-5 , max(dev[1:2,])+10 ) , xlim=c(1,5.1), xlab="number of parameters" , ylab="deviance" , pch=16 , col=rangi2)
mtext( concat( "N = ",N ) )
points( (1:5)+0.1 , dev[2,] )
for ( i in kseq ) {
  pts_in <- dev[1,i] + c(-1,+1)*dev[3,i]
  pts_out <- dev[2,i] + c(-1,+1)*dev[4,i]
  lines( c(i,i) , pts_in , col=rangi2 )
  lines( c(i,i)+0.1 , pts_out )
}

```

## Fit vs. Prediction

Start with a skeptical (**regularizing**) prior that makes the data work harder for the update.

## Assigned Problems

### Easy Problems

#### 7E1.  State the three motivating criteria that define information entropy. Try to express each in your own words.

(1) Continuity. The measure is not discrete, but it may be bounded. So it can have a minimum and maximum, but it must be continuous in between.
(2) Increasing with the number of events. When more distinct things can happen, and all else is equal, there is more uncertainty than when fewer things can happen.
(3) Additivity. This is hardest one to understand, for most people. Additivity is desirable because it means we can redefine event categories without changing the amount of uncertainty.

#### 7E2.  Suppose a coin is weighted such that, when it is tossed and lands on a table, it comes up heads 70% of the time. What is the entropy of this coin?


```{r 7E2}
# define probabilities of heads and tails
p <- c( 0.7 , 0.3 )
# compute entropy
-sum( p*log(p) )
```

#### 7E3.  Suppose a four-sided die is loaded such that, when tossed onto a table, it shows “1” 20%, “2” 25%, “3” 25%, and “4” 30% of the time. What is the entropy of this die?

```{r 7E3}
# define probabilities of sides
p <- c( 0.2 , 0.25 , 0.25 , 0.3 )
# compute entropy
-sum( p*log(p) )
```


#### 7E4.  Suppose another four-sided die is loaded such that it never shows “4”. The other three sides show equally often. What is the entropy of this die?

```{r 7E4}
# define probabilities of sides
p <- c( 1/3 , 1/3 , 1/3 )
# compute entropy
-sum( p*log(p) )
```


### Medium Problems

#### 7M1.  Write down and compare the definitions of AIC and WAIC. Which of these criteria is most general? Which assumptions are required to transform the more general criterion into a less general one?

AIC is...

$$AIC = -2log\;P(data|MAP est) + 2p = D_{train} + 2p$$
where $p$ is hte number of parameters in the model.  WAIC is tougher by notation but is...

$$WAIC = -2(lppd -p_{WAIC}) = -2(\sum log\;P(y_i) - \sum V(y_i)) $$
where $P(y_i)$ is the likelihood of observation $i$, averaging over the posterior distribution, and $V(y_i)$ is the variance in the log-likelihood of observation i, taking the variance over the posterior distribution.
Comparing these definitions, each has a term that expresses the fit to the training sample, as well as a term that expresses some penalty for flexibility in fitting to the sample. For AIC, the fit term is simply the "plug-in deviance," $D_{train}$, and the penalty term is twice the number of parameters, $p$.  For WAIC, the fit term is a fully Bayesian log-likelihood that averages over the posterior prediction for each observation separately. This is lppd. The penalty term is the sum of the variances of each log-likelihood.

From more general to less general: WAIC, AIC. When the posterior predictive mean is a good representation of the posterior predictive distribution, and the priors are effectively flat or overwhelmed by the amount of data, then WAIC and AIC will tend to agree.

#### 7M2.  Explain the difference between model selection and model comparison. What information is lost under model selection?

Model selection means ranking models by information criteria (or any other criteria) and choosing the highest ranked model. Model averaging is instead weighting each model by its relative distance from the best model and creating a posterior predictive distribution comprising predictions from each model in proportion to these weights, a prediction “ensemble.” Model selection discards information about model uncertainty. Model averaging retains some of that information, but it still discards some information about model uncertainty, because it compresses the full information about the set of models into a single predictive distribution. Since the full set of ranks and information criteria values cannot be recovered from the averaged predictive distribution, some of that information has been lost. In other words, different model comparison sets with different ranks and information criteria values can produce nearly identical prediction ensembles. But those different model comparison sets may be different in other important ways. For example, inspecting the full model comparison set, with the structure of each model, often reveals why some models fit better than others.

#### 7M3.  When comparing models with an information criterion, why must all models be fit to exactly the same observations? What would happen to the information criterion values, if the models were fit to different numbers of observations? Perform some experiments, if you are not sure.

When one model is fit to fewer (or different) observations, it is being judged on a different target than the other models. If fewer observations are used, the model will usually appear to perform better, because the deviance will smaller due to having less to predict. Less to predict means less prediction error, and deviance (as well as information criteria) is a kind of accumulated error. We have to be careful about this, because most of R’s black-box regression functions like lm, glm, and glmer will automatically and silently drop incomplete cases, reducing the number of observations the model is fit to. This is bad behavior for scientific software, but unfortunately it is the norm.

#### 7M4.  What happens to the effective number of parameters, as measured by PSIS or WAIC, as a prior becomes more concentrated? Why? Perform some experiments, if you are not sure.

As a prior becomes more concentrated around particular parameter values, the model becomes less flexible in fitting the sample. One way to remember this is to think of the prior as representing previous learning from previous observations. So a more concentrated, or peaked, prior represents more previous data. As the model becomes less flexible, the effective number of parameters declines. To perform some simple experiments to demonstrate this, try out this code, changing the value for sigma in the data list. The smaller sigma, the more concentrated the prior and the smaller the
effective number of parameters should be.


```{r 7M4}
y <- rnorm(10) # execute just once, to get data
# repeat this, changing sigma each time
m <- quap(
  alist(
    y ~ dnorm(mu,1),
    mu ~ dnorm(0,sigma)
    ), data=list(y=y,sigma=10) )
WAIC(m)
```

#### 7M5.  Provide an informal explanation of why informative priors reduce overfitting.

Informative priors reduce overfitting by reducing the sensitivity of a model to a sample. Some of the information in a sample is irregular, not a recurring feature of the process of interest.

#### 7M6.  Provide an informal explanation of why overly informative priors result in underfitting.

If a prior is overly informative, then even regular features of a sample will not be learned by a model. In this case, the model may underfit the sample. In case this sounds like a terrifying balancing act, in practice there is usually a broad family of priors which achieve practically indistinguishable estimates from the same sample. The last “hard” practice problem for this chapter provides an example.

### Hard Problems

#### 7H1.  In 2007, The Wall Street Journal published an editorial (“We’re Number One, Alas”) with a graph of corporate tax rates in 29 countries plotted against tax revenue. A badly fit curve was drawn in (reconstructed at right), seemingly by hand, to make the argument that the relationship between tax rate and tax revenue increases and then declines, such that higher tax rates can actually produce less tax revenue. I want you to actually fit a curve to these data, found in data(Laffer). Consider models that use tax rate to predict tax revenue. Compare, using WAIC or PSIS, a straight-line model to any curved models you like. What do you conclude about the relationship between tax rate and tax revenue?

#### 7H2.  In the Laffer data, there is one country with a high tax revenue that is an outlier. Use PSIS and WAIC to measure the importance of this outlier in the models you fit in the previous problem. Then use robust regression with a Student’s t distribution to revisit the curve fitting problem. How much does a curved relationship depend upon the outlier point?

It’s probably obvious from the prompt that the original curve was hand-drawn and not
the result of any fitting procedure. And you can see there is a high outlier point, while the rest of the points show a general increase. But let’s take this seriously and polish our modeling skills. Lots of combinations of models will produce the same general inference. I’ll start with comparing linear to polynomial models. Then I’ll try a spline. Then I’ll consider a robust regression to cope better with the outlier.

Here is a linear fit, together with a quadratic and cubic fit:

```{r 7H1and2}
library(rethinking)
data(Laffer)
d <- Laffer
d$T <- standardize( d$tax_rate )
d$R <- standardize( d$tax_revenue )
# linear model
m7H1a <- quap(
  alist(
    R ~ dnorm( mu , sigma ),
    mu <- a + b*T,
    a ~ dnorm( 0 , 0.2 ),
    b ~ dnorm( 0 , 0.5 ),
    sigma ~ dexp(1)
    ) , data=d )
# quadratic model
m7H1b <- quap(
  alist(
    R ~ dnorm( mu , sigma ),
    mu <- a + b*T + b2*T^2,
    a ~ dnorm( 0 , 0.2 ),
    c(b,b2) ~ dnorm( 0 , 0.5 ),
    sigma ~ dexp(1)
    ) , data=d )
# cubic model
m7H1c <- quap(
  alist(
    R ~ dnorm( mu , sigma ),
    mu <- a + b*T + b2*T^2 + b3*T^3,
    a ~ dnorm( 0 , 0.2 ),
    c(b,b2,b3) ~ dnorm( 0 , 0.5 ),
    sigma ~ dexp(1)
    ) , data=d )
compare( m7H1a , m7H1b , m7H1c , func=PSIS )
```

How do these models compare?

Not much support for the wiggly models over the linear one. But PSIS also warns about high leverage points. You can probably guess from scatterplot which point is causing the problem. If you peek at the pointwise k values, you’ll see:

```{r 7H1and2c1}
PSISk(m7H1a)
```

Point 12 is much over 1. That’s the nation with the very high tax revenue value. I think it’s Norway, actually, and it results from an accounting trick involving oil revenue. It’ll be useful to plot the posterior predictions of each model.

```{r 7H1and2c2}
T_seq <- seq( from=-3.2 , to=1.2 , length.out=30 )
la <- link( m7H1a , data=list(T=T_seq) )
lb <- link( m7H1b , data=list(T=T_seq) )
lc <- link( m7H1c , data=list(T=T_seq) )
plot( d$T , d$R , xlab="tax rate" , ylab="revenue" )
mtext( "linear model" )
lines( T_seq , colMeans(la) )
shade( apply( la , 2 , PI ) , T_seq )
plot( d$T , d$R , xlab="tax rate" , ylab="revenue" )
mtext( "quadratic model" )
lines( T_seq , colMeans(lb) )
shade( apply( lb , 2 , PI ) , T_seq )
plot( d$T , d$R , xlab="tax rate" , ylab="revenue" )
mtext( "cubic model" )
lines( T_seq , colMeans(lc) )
shade( apply( lc , 2 , PI ) , T_seq )
```

The polynomial models do bend the right way, but they aren’t bending much. If anything, they offer stronger evidence that more tax produces more revenue, not that very high tax reduces revenue. For the sake of the exercise, let’s consider a basis spline. Refer back to Chapter 4, the last section, if you’ve forgotten splines.

```{r 7H1and2c3}
num_knots <- 15
knot_list <- quantile( d$T , probs=seq(0,1,length.out=num_knots) )
library(splines)
B <- bs(d$T,
knots=knot_list[-c(1,num_knots)] ,
degree=3 , intercept=TRUE )
m7H1s <- quap(
  alist(
    R ~ dnorm( mu , sigma ) ,
    mu <- a + B %*% w ,
    a ~ dnorm(0,1),
    w ~ dnorm(0,1),
    sigma ~ dexp(1)
    ), data=list( R=d$R , B=B ) ,
  start=list( w=rep( 0 , ncol(B) ) ) )
compare( m7H1a , m7H1b , m7H1c , m7H1s , func=PSIS )
```

The coefficients aren’t interpretable, but let’s compare to the previous models (see above).  Not good.  What do the predictions look like?

```{r 7H1and2c4}
mu <- link( m7H1s )
mu_PI <- apply( mu , 2 , PI )
plot( d$T , d$R , xlab="tax rate" , ylab="revenue" )
mtext( "basis spline" )
o <- order( d$T )
lines( d$T[o] , colMeans(mu)[o] )
shade( mu_PI[,o] , d$T[o] )
```

If you really want to interpret that as support for the curved relationship, good luck. You might try more rigid splines, with lower polynomial order, fewer knots, and tighter weights, to see if this can be made more sensible. But clearly this is a curve-fitting exercise, not science, at this point. For the Student-t model, we just need to replace the normal likelihood. I’ll use ν = 2, so the tails
are satisfyingly thick.

```{r 7H1and2c5}
# linear model with student-t
m7H1d <- quap(
alist(
  R ~ dstudent( 2 , mu , sigma ),
  mu <- a + b*T,
  a ~ dnorm( 0 , 0.2 ),
  b ~ dnorm( 0 , 0.5 ),
  sigma ~ dexp(1)
  ) , data=d )
compare( m7H1a , m7H1b , m7H1c , m7H1s , m7H1d , func=PSIS )
```
Oh PSIS likes this one. It also doesn’t itself throw any Pareto-k warnings. If you like, try to polynomial models with the Student-t likelihood. Overall, there is no obvious support for a strongly increasingthen-decreasing relationship between tax rate and tax revenue.

#### 7H3.  Consider three fictional Polynesian islands. On each there is a Royal Ornithologist charged by the king with surveying the bird population. They have each found the following proportions of 5 important bird species:

<data omitted>

Notice that each row sums to 1, all the birds. This problem has two parts. It is not computationally complicated. But it is conceptually tricky. First, compute the entropy of each island’s bird distribution. Interpret these entropy values. Second, use each island’s bird distribution to predict the other two. This means to compute the KL divergence of each island from the others, treating each island as if it were a statistical model of the other islands. You should end up with 6 different KL divergence values. Which island predicts the others best? Why?

To compute the entropies, we just need a function to compute the entropy. Information entropy, as defined in lecture and the book, is simply:

$$H(p) = -\sum_i p_i log(p_i) $$
where p is a vector of probabilities summing to 1.  In R, we code the function like this:

```{r 7H3}
H <- function(p){
  out <- -sum(p*log(p))
  return(out)
}
```

Here are the data and application to the function H:

```{r 7H3c1}
IB <- list()
IB[[1]] <- c( 0.2 , 0.2 , 0.2 , 0.2 , 0.2 )
IB[[2]] <- c( 0.8 , 0.1 , 0.05 , 0.025 , 0.025 )
IB[[3]] <- c( 0.05 , 0.15 , 0.7 , 0.05 , 0.05 )
sapply( IB , H )
```

The first island has the largest entropy, followed by the third, and then the second in last place. Why is this? Entropy is a measure of the evenness of a distribution. The first islands has the most even distribution of birbs. This means you wouldn’t be very surprised by any particular bird. The second island, in contrast, has a very uneven distribution of birds. If you saw any bird other than the first species, it would be surprising.

Now we need K-L distance, so let’s write a function for it:

```{r 7H3c2}
DKL <- function(p,q){
  out <- sum( p*(log(p)-log(q)) )
  return(out)
}
```

This is the distance from q to p, regarding p as true and q as the model. Now to use each island as a model of the others, we need to consider the different ordered pairings. I’ll just make a matrix and loop over rows and columns:

```{r 7H3c3}
Dm <- matrix( NA , nrow=3 , ncol=3 )
for ( i in 1:3 ){
  for ( j in 1:3 ){
    Dm[i,j] <- DKL( IB[[j]] , IB[[i]] )
  }
}
round( Dm , 2 )
```

The way to read this is each row as a model and each column as a true distribution. So the first island, the first row, has the smaller distances to the other islands. This makes sense, since it has the highest entropy. Why does that give it a shorter distance to the other islands? Because it is less surprised by the other islands, due to its high entropy


#### 7H4.  Recall the marriage, age, and happiness collider bias example from Chapter 6. Run models m6.9 and m6.10 again (page 178). Compare these two models using WAIC (or PSIS, they will produce identical results). Which model is expected to make better predictions? Which model provides the correct causal inference about the influence of age on happiness? Can you explain why the answers to these two questions disagree?

```{r 7H4}
library(rethinking)
d <- sim_happiness( seed=1977 , N_years=1000 )
d2 <- d[ d$age>17 , ] # only adults
d2$A <- ( d2$age - 18 ) / ( 65 - 18 )
d2$mid <- d2$married + 1
m6.9 <- quap(
  alist(
    happiness ~ dnorm( mu , sigma ),
    mu <- a[mid] + bA*A,
    a[mid] ~ dnorm( 0 , 1 ) ,
    bA ~ dnorm( 0 , 2 ) ,
    sigma ~ dexp(1)
    ) , data=d2 )

m6.10 <- quap(
  alist(
    happiness ~ dnorm( mu , sigma ),
    mu <- a + bA*A,
    a ~ dnorm( 0 , 1 ),
    bA ~ dnorm( 0 , 2 ),
    sigma ~ dexp(1)
    ) , data=d2 )

compare( m6.9 , m6.10 )
```

The model that produces the invalid inference, m6.9, is expected to predict much better. And it would. This is because the collider path does convey actual association. We simply end up mistaken about the causal inference. We should not use WAIC (or LOO) to choose among models, unless we have some clear sense of the causal model. These criteria will happily favor confounded models.

#### 7H5.  Revisit the urban fox data, data(foxes), from the previous chapter’s practice problems. Use WAIC or PSIS based model comparison on five different models, each using weight as the outcome, and containing these sets of predictor variables:
(1)  avgfood + groupsize + area
(2)  avgfood + groupsize
(3)  groupsize + area
(4)  avgfood
(5)  area
Can you explain the relative differences in WAIC scores, using the fox DAG from the previous chapter? Be sure to pay attention to the standard error of the score differences (dSE).

```{r 7H5}
library(rethinking)
data(foxes)
d <- foxes
d$W <- standardize(d$weight)
d$A <- standardize(d$area)
d$F <- standardize(d$avgfood)
d$G <- standardize(d$groupsize)

m1 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bF*F + bG*G + bA*A,
    a ~ dnorm(0,0.2),
    c(bF,bG,bA) ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )

m2 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bF*F + bG*G,
    a ~ dnorm(0,0.2),
    c(bF,bG) ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )

m3 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bG*G + bA*A,
    a ~ dnorm(0,0.2),
    c(bG,bA) ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )

m4 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bF*F,
    a ~ dnorm(0,0.2),
    bF ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )

m5 <- quap(
  alist(
    W ~ dnorm( mu , sigma ),
    mu <- a + bA*A,
    a ~ dnorm(0,0.2),
    bA ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    ), data=d )

compare( m1 , m2 , m3 , m4 , m5 )
```

Notice that the top three models are m1, m3, and m2. They have very similar WAIC values. The differences are small and smaller in all cases than the standard error of the difference. WAIC sees these models are tied. This makes sense, given the DAG, because as long as a model has groupsize in it, we can include either avgfood or area or both and get the same inferences. Another way to think of this is that the influence of good, adjusting for group size, is (according to the DAG) the same as the influence of area, adjusting for group size, because the influence of area is routed entirely through food and group size. There are no backdoor paths.

What about the other two models, m4 and m5? These models are tied with one another, and both omit group size. Again, the influence of area passes entirely through food. So including only food or only area should produce the same inference—the total causal influence of area (or food) is just about zero. That’s indeed what the posterior distributions suggest:

```{r 7H5c1}
coeftab(m4,m5)
```

